---
search: false
format: rostools-theme-revealjs
---

# Welcome to the Advanced R^3^ workshop!

-   :heavy_check_mark: Pick up a table name from the bowl
-   :heavy_check_mark: Go to the table name you picked
-   :heavy_check_mark: Introduce yourself to your table mates
-   :heavy_check_mark: Open your `AdvancedR3.Rproj` RStudio project

::: notes
Introduce yourself and instructors.
:::

# Motivation for this workshop

## Largely from desire for more effective and actual collaboration {.center}

::: notes
Relate to UK Biobank project: <https://github.com/steno-aarhus/ukbAid>
:::

## But also from larger trends in research and society {.center}

### Like reproducibility, "AI", and team science

::: notes
Ask who knows what reproducibility is and could explain it.
:::

## Very low reproducibility in most of science [@Samuel2024] {.smaller}

::: {layout-ncol="2"}
```{r}
#| echo: false
#| fig-height: 6
library(tidyverse)
theme_set(
  theme_minimal() +
    theme(
      plot.background = element_rect(fill = "#E9E9E9"),
      axis.title = element_blank(),
      axis.text.y = element_text(size = 20),
      axis.text.x = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank()
    )
)

search <- tribble(
  ~item, ~description, ~value, ~order,
  "Articles", "Found among all articles", 3467, 1,
  "GitHub\nrepos", "Those with a GitHub link and at least one notebook", 2660, 2
) |>
  mutate(
    item = fct_rev(fct_reorder(item, order))
  )

ggplot(search, aes(y = item, x = value, label = value)) +
  geom_col(fill = "#203C6E") +
  geom_text(nudge_x = 200, size = 6)
```

```{r}
#| echo: false
#| fig-height: 6
repro_results <- tribble(
  ~item, ~description, ~value, ~order,
  "Total notebooks", "Total notebook files found in all the repositories", 27271, 1,
  "Notebook repo\nwas valid", "File names were correct, dependency information was available, no local modules were needed.", 15817, 2,
  "Could install\ndependencies", "Could install dependencies without errors.", 10388, 3,
  "Finished executing\nwithout errors", "Could run without any errors.", 1203, 4,
  "Same results\nas paper", "When the results of the paper matched what the execution results output.", 879, 5
) |>
  mutate(
    percent = str_c(value, "\n(", round(value / max(value) * 100), "%)"),
    item = fct_rev(fct_reorder(item, order))
  )

ggplot(repro_results, aes(y = item, x = value, label = percent)) +
  geom_col(fill = "#203C6E") +
  geom_text(nudge_x = 1700, size = 6)
```
:::

From searching PubMed Central for all articles with search term: "(ipynb
OR jupyter OR ipython) AND github"

## Researchers don't know how to collaborate effectively {.center}

We don't get training in it, nor do we get much practice in it

::: notes
You do your PhD alone, you get grants as a postdoc for yourself, etc.
:::

## Academia doesn't encourage building and improving effective collaboration {.center}

::: notes
Getting grants means getting publications, getting publications means
being involved in lots of project, being involved in lots of projects
means contributing mainly through comments or feedback or meetings
(since you only have so much time in a day), getting funding means you
need to get yourself as last author on as many papers as possible to
continue getting funding, etc.

It's a vicious cycle.
:::

## One direct way to collaborate is working on shared code for a project {.center}

## Reading *others'* code is necessary to improve as an experienced coder {.smaller}

-   You aren't reading others' code if:

    -   you don't collaborate with others *directly* on their code
    -   you aren't doing code reviews with colleagues
    -   researcher's don't share their code

. . .

-   If *we all* aren't reading others' code, *we all* aren't improving

## As you get more advanced, you learn that more coding isn't a good thing {.center}

Research "product" is an *asset*, but *code* is a *liability*

::: notes
As you get more advanced in coding and analysis, you actually spend less
time coding and more time thinking about the analysis, the design of the
analysis and the code, and the overall flow of things. You start
learning that, while your finished analysis is an *asset*, the more
lines of code you have for it is a *liability*. You want your code to be
as simple and minimal as possible, while still doing what you need it to
do.
:::

## With "AI" tools, generating (or "vibing") lots of code is easy... but... {.center}

## Problem: If you aren't experienced in code, how can you assess it? {.center}

::: notes
Problem with AI tools, it predicts the average next items. Key word,
average. Do you want to go for average? Is average code even good?
There's a lot of shit code in the world, that, the AI has been trained
on.
:::

## More code = means more reading, maintenance, and hidden bugs {.center}

## Designing the analysis and code becomes even more important {.center}

Even before looking at the data

::: notes
This is something that LLMs or machine learning algorithms can't do, is
understand human context or needs or wants or intentions or aims. Hell,
communicating to other humans to get everyone on the same page on its
own is incredibly hard.
:::

## Well-designed code is a powerful enabler of effective collaboration {.center}

-   Better at communicating intent

-   Does more, with less

::: notes
You can do a lot with surprisingly little code if you design it well.
That's part of what we want you to learn from this workshop: do more
with less.
:::

## Goal of workshop: Create and nurture ... {.center}

-   A culture of openness

-   Design thinking

-   Effective collaboration

::: notes
In this workshop, we'll be focusing more on the bigger picture and
design, and in fact, probably do less coding than in previous workshops.
We'll do more thinking first, then design, then code.
:::

# Workshop details

## Setup and layout {.smaller}

::::: columns
::: {.column width="50%"}
-   Workshop is mix of:
    -   "Code-alongs" (we type and explain, you type along)
    -   Hands-on coding, discussing, and reading exercises
    -   A group project at the end
-   All material is online (and [openly
    licensed](https://r-cubed-advanced.rostools.org/LICENSE.html))
:::

::: {.column width="50%"}
-   Less about coding, more about:
    -   Connecting and collaborating with others
    -   Thinking about designing first, coding second
-   Networking and connecting:
    -   Draw table name from bowl each morning
    -   Introduce yourself to those at your table
    -   During lunch, sit beside someone you don't know
    -   Several networking activities (usually after lunch)
:::
:::::

## Getting or asking for help :raising_hand_woman:

::::: columns
::: {.column width="50%"}
-   Put the sticky (a.k.a. tiny paper hat :tophat:) on your laptop to
    get help
-   There are lots of helpers
-   Table mates, help out too
:::

::: {.column width="50%"}
-   We're all learning here!
-   This is a supportive and safe environment
-   Remember our **Code of Conduct**
:::
:::::

## Practice using stickies/paper hats: Do you have RStudio open? {.center}

# Activities

## :walking_woman::walking_man: How do you perceive your skill in R? {.center}

::: notes
Along the wall, arrange to one side is "novice/basic" and other side is
"advanced".
:::

## :walking_woman::walking_man: If code for a paper was audited, how confident are you that the results could be reproduced *exactly*? {.center}

No judgment!

::: notes
Range from not confident to very confident.
:::

## :walking_woman::walking_man: How often do you intentionally design your analysis before doing it? {.center}

::: notes
Range from never to always.
:::

## :walking_woman::walking_man: How often do you work on your own in your project(s)? {.center}

::: notes
Range from almost entirely alone to never alone.
:::

## :walking_woman::walking_man: How easy is it for others to collaborate on your project(s)? {.center}

::: notes
Range from not easy to very easy.
:::

## :walking_woman::walking_man: How many of you want to work more closely or collaboratively with others? {.center}

::: notes
Range from definitely no to definitely yes.
:::

## Alright, return to your seats! :chair: {.center}
