---
search: false
format: r3-theme-revealjs
bibliography: "../includes/references.bib"
csl: "../_extensions/rostools/r3-theme/vancouver.csl"
---

# Welcome to the Advanced R^3^ course!

-   :heavy_check_mark: Go to your assigned table
-   :heavy_check_mark: Introduce yourselves to your table mates
-   :heavy_check_mark: Open your `AdvancedR3.Rproj` RStudio project
-   :heavy_check_mark: Just in case, do
    `r3::install_packages_advanced()`

::: notes
Introduce yourself and instructors.
:::

# Motivation for this course

## Largely from desire for more effective and actual collaboration {.center}

::: notes
Relate to UK Biobank project: <https://github.com/steno-aarhus/ukbAid>
:::

## But also from larger trends in research and society {.center}

### Like reproducibility and more team science

::: notes
Ask who knows what reproducibility is and could explain it.
:::

## Very low reproducibility in most of science [@Samuel2024] {.smaller}

::: {layout-ncol="2"}
```{r}
#| echo: false
#| fig-height: 6
library(tidyverse)
theme_set(
  theme_minimal() +
    theme(
      plot.background = element_rect(fill = "#E9E9E9"),
      axis.title = element_blank(),
      axis.text.y = element_text(size = 20),
      axis.text.x = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank()
    )
)

search <- tribble(
  ~item, ~description, ~value, ~order,
  "Articles", "Found among all articles", 3467, 1,
  "GitHub\nrepos", "Those with a GitHub link and at least one notebook", 2660, 2
) |>
  mutate(
    item = fct_rev(fct_reorder(item, order))
  )

ggplot(search, aes(y = item, x = value, label = value)) +
  geom_col(fill = "#203C6E") +
  geom_text(nudge_x = 200, size = 6)
```

```{r}
#| echo: false
#| fig-height: 6
repro_results <- tribble(
  ~item, ~description, ~value, ~order,
  "Total notebooks", "Total notebook files found in all the repositories", 27271, 1,
  "Notebook repo\nwas valid", "File names were correct, dependency information was available, no local modules were needed.", 15817, 2,
  "Could install\ndependencies", "Could install dependencies without errors.", 10388, 3,
  "Finished executing\nwithout errors", "Could run without any errors.", 1203, 4,
  "Same results\nas paper", "When the results of the paper matched what the execution results output.", 879, 5
) |>
  mutate(
    percent = str_c(value, "\n(", round(value / max(value) * 100), "%)"),
    item = fct_rev(fct_reorder(item, order))
  )

ggplot(repro_results, aes(y = item, x = value, label = percent)) +
  geom_col(fill = "#203C6E") +
  geom_text(nudge_x = 1700, size = 6)
```
:::

From searching PubMed Central for all articles with search term: "(ipynb
OR jupyter OR ipython) AND github"

## Researchers don't know how to collaborate effectively {.center}

We don't get training in it, nor do we get much practice in it

::: notes
You do your PhD alone, you get grants as a postdoc for yourself, etc.
:::

## Academia doesn't encourage building and improving effective collaboration {.center}

::: notes
Getting grants means getting publications, getting publications means
being involved in lots of project, being involved in lots of projects
means contributing mainly through comments or feedback or meetings
(since you only have so much time in a day), getting funding means you
need to get yourself as last author on as many papers as possible to
continue getting funding, etc.

It's a vicious cycle.
:::

## Little published work on how to effectively collaborate {.center}

An example [@Turoman2022]: Research group tries to work on improving
their collaboration

-   Decide on tools and practices that are best for the group
-   Use "data partners" to co-work/co-code together

::: notes
I'm collaborating with some colleagues on a scoping review for open
collaboration, and there is very little research in this area, or even
"guides" or "examples", at least not found in literature databases.
:::

## One direct way to collaborate is working on shared code for a project {.center}

Because by collaborating, you will also read others' code.

## Reading *others'* code is necessary to improve as an experienced coder {.smaller}

-   You aren't reading others' code if:

    -   you don't collaborate with others *directly* on their code
    -   you aren't doing code reviews with colleagues
    -   researcher's don't share their code

. . .

-   If *we all* aren't reading others' code, *we all* aren't improving

## Goal of course: Create and nurture a culture of openness through training {.center}

# Course details

## Setup and layout {.smaller}

::::: columns
::: {.column width="50%"}
-   Course is mix of:
    -   "Code-alongs" (we type and explain, you type along)
    -   Hands-on coding, discussing, and reading exercises
    -   A group project at the end
-   All material is online (and [openly
    licensed](https://r-cubed-advanced.rostools.org/LICENSE.html))
:::

::: {.column width="50%"}
-   Less about coding, more about connecting with others
    -   Draw table name from bowl each morning
    -   Introduce yourself to those at your table
    -   During lunch, sit beside someone you don't know
    -   Several networking activities (usually after lunch)
-   *Warning*: You will be making a lot of functions!
:::
:::::

## Getting or asking for help :raising_hand_woman:

::::: columns
::: {.column width="50%"}
-   Put the sticky (a.k.a. tiny paper hat :tophat:) on your laptop to
    get help
-   There are lots of helpers
-   Table mates, help out too
:::

::: {.column width="50%"}
-   We're all learning here!
-   This is a supportive and safe environment
-   Remember our **Code of Conduct**
:::
:::::

## Practice using stickies/paper hats: Did you complete the checklist from the beginning? {.center}

# Activities

## :walking_woman::walking_man: How do you perceive your skill in R? {.center}

::: notes
Along the wall, arrange to one side is "novice/basic" and other side is
"advanced".
:::

## :raising_hand_woman: Who has had formal training in specifically *coding* in R or in general? {.center}

::: notes
Staying where you are, raise your hand if you would answer yes to this
question. If it was part of a statistics course, it doesn't really
count.
:::

## :walking_woman::walking_man: Have you attended at least one of these courses before? {.center}

::: notes
Only have those who indicated they did have training (from previous
question).
:::

## :walking_woman::walking_man: If code for a paper was audited, how confident are you that the results could be reproduced *exactly*? {.center}

::: notes
Range from not confident to very confident.
:::

## :walking_woman::walking_man: How many of you work mostly on your own in your project(s)? {.center}

::: notes
Range from almost entirely alone to never alone.
:::

## :walking_woman::walking_man: How easy is it to collaborate with others on your project(s)? {.center}

::: notes
Range from not easy to very easy.
:::

## :walking_woman::walking_man: How many of you want to work more closely or collaboratively with others? {.center}

::: notes
Range from definitely no to definitely yes.
:::

## Alright, return to your seats! :chair: {.center}
