# Setting up automatic analysis pipelines {#sec-pipelines-setup}

```{r setup}
#| include: false
source(here::here("R/functions.R"))
library(tidyverse)
library(targets)
lipidomics <- read_csv(here::here("data/lipidomics.csv"))
```

{{< text_snippet review_note >}}

When doing analyses on data, you have probably experienced (many) times
where you have to re-run some analyses and have forgotten the order that
code should run or which parts you need to re-run in order to update the
results. Things can get confusing quickly, even in relatively simple
projects. This becomes even more challenging when you return to a
project after a month or two and have forgotten the state of the
analysis and the project as a whole. Add in collaborators, and things
can get even more complex.

That's where formal data analysis pipeline tools come in. By organising
your analyses into distinct steps, with clear inputs and outputs, and
adding these steps to a pipeline that tracks them, you can make things a
lot easier for yourself and others. This session focuses on using tools
that create and manage these pipelines effectively.

## Learning objectives

{{< include /includes/objectives/_pipelines-setup.qmd >}}

## :speech_balloon: Discussion activity: How do you re-run analyses when something changes?

**Time: \~12 minutes.**

We've all been (or will be) in situations where something in our
analysis needs to change: Maybe we forgot to remove a certain condition
(like unrealistic BMI values), maybe our supervisor suggests something
we hadn't considered in our analysis, or maybe during peer review of our
manuscript, a reviewer makes a suggestion that would improve the paper.

Whatever the situation, we inevitably need to re-run parts of or the
full analysis. So what is your **exact** workflow when you need to
re-run code and update your results? Assume it's a change in the code
somewhere early in the data processing stage.

1.  Take about 1 minute to think about the workflow you use. Try to
    think of the *exact* steps you need to take, what *exactly* you do,
    and how long that usually takes.
2.  For 8 minutes, share and discuss your thoughts in with your
    neighbour. How do your experiences compare to each other?
3.  For the remaining time, we'll briefly share with everyone what
    they've thought and discussed.

## :book: Reading task: What is a data analysis "pipeline"? {#sec-what-is-pipeline}

**Time: \~10 minutes.**

::: {.callout-note collapse="true"}
## :teacher: Instructor note

After they finish reading this section, briefly walk through it. In
particular, emphasize what we want to make at the end, even though that
goal might change as the analysis project progresses.
:::

A pipeline can be any process where the steps between a start and an end
point are very clear, explicit, and concrete. These highly distinct
steps can be manual, human involved, or completely automated by a robot
or computer. For instance, in car factories, the pipeline from the input
raw materials to the final vehicle is extremely well described and
implemented. Similarly, during the pandemic, the pipeline for testing
(at least in Denmark and several other countries) was highly structured
and clear for both the workers doing the testing and the people having
the test done: A person goes in, scans their ID card, has the test done,
the worker inputs the results, and the results are sent immediately to
the health agency as well as to the person based on their ID contact
information (or via a secure app).

However, in research, especially around data collection and analysis,
you may often hear or read about "pipelines". But looking closer, these
aren't actual pipelines because the individual steps are not very clear
and not well described, and they often require a fair amount of manual
human attention and intervention. Often these "pipelines" are general
workflows that people follow to complete a specific set of tasks.

In computational environments, a pipeline should be a set of data
processing steps connected in a series with a specific order; the output
of one step is the input to the next. This means that there actually
should be minimal to *no* human intervention from raw input to finished
product. And it means that the entire process should be *automated* and
*reproducible*.

So why are these computational data pipelines rarely found in most of
the research "pipelines" described in papers? We can break the reason
why not into specific points that together explain why not:

1.  Anything with data ultimately must be on or done by the computer,
2.  Anything automatically done on the computer must be done with code,
3.  Not all researchers write code,
4.  Researchers who do write code rarely publish and share it,
5.  Code that is shared or published (either publicly or within the
    research group) is not written in a way that allows a pipeline to
    exist,
6.  And, research is largely non-reproducible [@Trisovic2022;
    @Seibold2021; @Laurinavichyute2021].

A data analysis pipeline would, by definition, be a readable and
reproducible data analysis. But based on the above points, researchers
(on average and as a group) don't have the technical expertise and
knowledge to actually implement data analysis pipelines.

This isn't to diminish the work of researchers, but is rather a basic
observation on the systemic, social, and structural environment
surrounding us. We as researchers are not necessarily trained in writing
code, nor is there a strong culture and incentive structure around
learning, sharing, reviewing, and improving code. Which also results in
use very rarely being allowed to get (or use) funds to hire people who
*are* trained and skilled in programmatic designing, thinking, and
coding. Otherwise workshops like this wouldn't need to exist
:woman_shrugging:

Before we get to what an actual data analysis pipeline looks like in
practice, we have to separate two things: *exploratory data analysis*
and *data analysis for the final paper*. In exploratory data analysis,
there will likely be a lot of manual, interactive steps involved that
may or may not need to be explicitly stated and included in the analysis
plan and pipeline. But for the final paper and the included results, you
generally have some basic first ideas of what you'll need. Let's list a
few items that you would want to do before the primary statistical
analysis of a research project:

1.  A table of some basic descriptive statistics of the study
    population, such as mean, standard deviation, or counts of basic
    discrete data (like treatment group). Usually this is a "table 1" in
    the paper.
2.  A figure showing the distribution of your main variables of
    interest. In this case, ours are the lipidomic variables. Like with
    the table above, this is usually "figure 1" or a similarly early
    figure in the paper.
3.  Lastly, the paper with the above table and figure included.

Let's break this down visually in @fig-pipeline-schematic:

```{mermaid}
%%| label: fig-pipeline-schematic
%%| fig-cap: The flow of data into specific 'targets' as the initial steps in a data analysis pipeline. The "?" are actions (functions) that need to happen but that we haven't yet defined.
%%| echo: false
graph TB
    data[Data<br>in a file] -->|?| tab_desc[Item 1:<br>Descriptive<br>statistics<br>table]
    tab_desc -->|insert| paper[Item 3:<br>Paper]
    data -->|?| plot_distrib[Item 2:<br>Continuous<br>variables<br>plot]
    plot_distrib -->|insert| paper
```

Here, there are three distinct items ("products") that you (might) want
to create from your data. Each "flow" from data to the desired product
has specific actions that need to happen. Now that you've seen how it is
conceptually drawn out, there are some initial tasks you can work
towards in your pipeline. In this session, we will build up the actions
piece-by-piece to go from data to final product.

{{< text_snippet sticky_up >}}

## Using targets to manage a pipeline

There are a few packages that help build pipelines in R, but the most
commonly used, well-designed, and maintained one is called `{targets}`.
This package allows you to explicitly set "targets" (objects like the
table or figure) you want to create as part of the pipeline. `{targets}`
will then track all your targets for you and tracks which output a
target is input to another target as well as which target need to be
updated when you make changes to your pipeline.

::: {.callout-note collapse="true"}
## :teacher: Instructor note

Ask participants, which do they think it is: a build dependency or a
workflow dependency. Because it is directly used to run analyses and
process the data, it would be a build dependency.
:::

First, we need to install and add `{targets}` to our dependencies. Since
`{targets}` is a build dependency, we'll add it to the `DESCRIPTION`
file with:

```{r add-targets-as-dep}
#| purl: true
#| eval: false
#| filename: Console
use_package("targets", "suggests")
```

Now that it's added as a dependency, let's set up our project to start
using `{targets}`! There's a handy helper function that does this for
us:

```{r use-targets}
#| purl: true
#| eval: false
#| filename: Console
targets::use_targets()
```

This command will create a `_targets.R` script file that we'll use to
define our analysis pipeline. Before we do that though, let's **commit
the file to the Git history** with {{< var keybind.git >}}.

::: callout-note
If you are using an older version of the `{targets}` package, you might
also find that running `targets::use_targets()` also have created a
`run.R` and `run.sh` file. The files are used for other situations (like
running on a Linux server) that we won't cover in this workshop. You can
safely delete them.
:::

## :book: Reading task: Inside the `_targets.R` file

**Time: \~8 minutes.**

::: {.callout-note collapse="true"}
## :teacher: Instructor note

Let them read it before going over it again to reinforce the
function-oriented workflows and how `{targets}` and the `tar_target()`
work. In particular emphasise the relationship between the `{targets}`
`list(tar_target(...), ...)`, the function-oriented workflow, and the
graph.
:::

For this reading task, you'll take a look at what `_targets.R` contains.
So, start by opening it on your computer.

::: {.callout-note collapse="true"}
## Click this to see the file contents

```{r}
#| echo: false
#| code-fold: true
fs::path_package("targets", "pipelines", "use_targets.R") |>
  readLines() |>
  cat(sep = "\n")
```
:::

Mostly, the`_targets.R` script contains comments and a basic set up to
get you started. But, notice the `tar_target()` function used at the end
of the script. There are two main arguments for it: `name` and
`command`. The way that `{targets}` works is similar to how you'd assign
the output of a function to an object, so:

``` r
object_name <- function_in_command(input_arguments)
```

Is the same as:

``` r
tar_target(
  name = object_name,
  command = function_in_command(input_arguments)
)
```

What this means is that `{targets}` follows a ["function-oriented"
workflow](https://books.ropensci.org/targets/functions.html#problems-with-script-based-workflows),
not a "script-based" workflow. What's the difference? In a
script-oriented workflow, each R file/script is run in a specific order.
As a result, you might end up with a "main" R file that has code that
looks like:

``` r
source("R/1-process-data.R")
source("R/2-basic-statistics.R")
source("R/3-create-plots.R")
```

Where it will run each of these R scripts in a specific order.
Meanwhile, in a function-oriented workflow, it might look more like:

``` r
source("R/functions.R")
raw_data <- load_raw_data("file/path/data.csv")
processed_data <- process_data(raw_data)
basic_stats <- calculate_basic_statistics(processed_data)
simple_plot <- create_plot(processed_data)
```

With the function-oriented workflow, each function takes an input and
contains all the code to create one result as its output. This could be,
for instance, a figure in a paper.

If you've taken the [intermediate R
workshop](https://r-cubed-intermediate.rostools.org/), you'll notice
that this function-oriented workflow is the workflow we covered in that
workshop. There are so many advantages to this type of workflow which is
why many powerful R packages are designed around making use of it.

If we take the same code as above and convert it into the `{targets}`
format, the end of `_targets.R` would like this:

```{r}
#| eval: false
list(
  # This part is special, as it tells targets that the
  # character string is a file path.
  tar_target(
    name = data_file,
    command = "file/path/data.csv",
    format = "file"
  ),
  tar_target(
    name = raw_data,
    command = load_raw_data(data_file)
  ),
  tar_target(
    name = processed_data,
    command = process_data(raw_data)
  ),
  tar_target(
    name = basic_stats,
    command = calculate_basic_statistics(processed_data)
  ),
  tar_target(
    name = simple_plot,
    command = create_plot(processed_data)
  )
)
```

So, each `tar_target()` is a step in the pipeline. The `command` is the
function call, while the `name` is the name of the output of the
function call.

In `{targets}`, in order for it to effectively track, or "watch",
changes to a file, we have to tell it that a target is a file by using
the `format = "file"` argument. This way, if the file changes,
`{targets}` knows that it needs to re-run any targets that depend on it.

This relationship between objects and functions (because of the
function-oriented workflow) can be represented exactly as a graph, with
nodes and arrows. Each node is an R object and each arrow is the
function that converts that object to another object. Since the
`{targets}` pipeline is structured using a function-oriented workflow,
the pipeline can also be visually represented as a graph, which we show
in @fig-targets-as-flow.

```{mermaid}
%%| label: fig-targets-as-flow
%%| fig-cap: Representing the targets pipeline structure as a flow diagram, where each "node" is an R object and each arrow is the function that converts the object to another object. We'll explain and build up this pipeline step by step.
%%| echo: false
graph TB
    data_file -->|"load_raw_data()"| raw_data
    raw_data -->|"process_data()"| processed_data
    processed_data -->|"calculate_basic_statistics()"| basic_stats
    processed_data -->|"create_plot()"| simple_plot
```

And in fact, `{targets}` has some functionality to visually generate a
graph of the targets and their connections, which we will cover later in
this session.

{{< text_snippet sticky_up >}}

## Building up our first target

Now that we've conceptually learned about how `{targets}` works, let's
take a big step back and first graphically design our own pipeline. So
we want to create the three items we mentioned before: a descriptive
statistics table, a distribution plot of the lipidomic variables, and
the final paper with these items included. We also have our data in a
CSV file. So there are at least five targets we need to create in our
pipeline. Let's draw it out:

```{mermaid}
%%| label: fig-our-pipeline-schematic
%%| fig-cap: The design and flow of how we want our data analysis pipeline to look like, with the distinct targets we want to create.
%%| echo: false
graph TB
    data_file["data_file<br>'data/lipidomics.csv'"] -->|"read_csv()"| lipidomics
    lipidomics -->|"create_table_descriptive_stats()"| table_descriptive_stats
    lipidomics -->|"create_plot_distributions()"| plot_distributions
    table_descriptive_stats -->|"tar_read()"| report
    plot_distributions -->|"tar_read()"| report
```

The very first target we want to create is to load in data from a file.
So let's add it as a step in the `{targets}` pipeline. Open up the
`_targets.R` file and go to the end of the file, where the `list()` and
`tar_target()` code is found. Delete all but one of the `tar_target()`
inside the `list()`.

To track the actual data file, we will need to create a pipeline target
that defines the location of our file by using the argument
`format = "file"`.

```{r targets-data-path}
#| eval: false
#| filename: "_targets.R"
list(
  tar_target(
    name = file,
    command = "data/lipidomics.csv",
    format = "file"
  )
)
```

Let's run `{targets}` to see what happens! You can either use
{{< var keybind.targets-make >}} or run this code in the Console:

```{r targets-make}
#| eval: false
#| filename: "Console"
targets::tar_make()
```

For the rest of the workshop we will use
{{< var keybind.targets-make >}} (the "foreground" one) rather than the
function in the Console.

After running it, you should see in one of RStudio's panes something
that looks like:

```
#> + file dispatched
#> ✔ file completed
```

This tells us that the target called `file` has been run. Since we only
have one target, that's all that `{targets}` could run.

We also see that a new folder has been created called `_targets/`.
Inside this folder `{targets}` keeps all of the output from running the
code in `_targets.R`. It comes with its own `.gitignore` file so that
you don't track all the files inside, as they don't need to be tracked.
Only the `_targets/meta/meta` is needed to be tracked in Git.

We can visualize our pipeline now too! This can be useful as you add
more and more targets. We will (likely) need to install an extra package
(done automatically):

```{r visualize-targets}
#| purl: true
#| eval: false
#| filename: Console
targets::tar_visnetwork()
```

We can also use {{< var keybind.targets-vis >}}, which is what we will
use for the rest of the workflow rather than the Console command.

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Walkthrough and explain the visualisation. Take your time moving things
around, zooming in and out, and clicking around.
:::

If you want to check which targets are outdated, we use:

```{r outdated-targets}
#| purl: true
#| eval: false
#| filename: Console
targets::tar_outdated()
```

Or by using {{< var keybind.targets-outdated >}}.

Now, let's actually load in the data as a target. We'll create a new
`tar_target()` that reads in the data using `readr::read_csv()`, using
the `file` target as input. In the `_targets.R` file at the `list()`
section at the bottom, add another `tar_target()` that looks like:

```{r targets-read-data}
#| eval: false
#| filename: "_targets.R"
list(
  tar_target(
    name = file,
    command = "data/lipidomics.csv",
    format = "file"
  ),
  tar_target(
    name = lipidomics,
    command = readr::read_csv(file, show_col_types = FALSE)
  )
)
```

We use `show_col_types = FALSE` to suppress the message explaining which
columns `{readr}` is reading into R. Now, let's run `{targets}` again
using {{< var keybind.targets-make >}}. After it runs, check the
visualisation of it with {{< var keybind.targets-vis >}}. You can now
see that there are two targets, with an arrow from `file` to
`lipidomics`.

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Walkthrough this again, this time showing the new target that was added
and the arrow between the targets.
:::

How can we access that new `lipidomics` data that `{targets}` created?
For that, we use the function `targets::tar_read()`. `{targets}` stores
the output of each target in its own internal storage system, which is
kept in the `_targets/` folder. So when we use `targets::tar_read()`, we
access any stored target items using syntax like we would with
`{dplyr}`, without quotes.

So in the Console, let's see how our data looks like by running:

```{r read-lipidomics-target}
#| eval: false
#| filename: "Console"
targets::tar_read(lipidomics)
```

```{r admin-read-lipidomics-target}
#| echo: false
readr::read_csv(here::here("data/lipidomics.csv"), show_col_types = FALSE)
```

We use this syntax whenever we want to output or print a target object,
including for tables and figures.

```{r purl-only-add-to-targets}
#| eval: false
#| echo: false
#| purl: true
new_targets <- '
list(
    tar_target(
        name = file,
        command = "data/lipidomics.csv",
        format = "file"
    ),
    tar_target(
        name = lipidomics,
        command = readr::read_csv(file, show_col_types = FALSE)
    )
)
'
# print_lines("_targets.R")
revise_by_line_num(
  path = "_targets.R",
  insert_text = new_targets,
  remove_original_lines = -28:-38,
  insert_at_line = 28
)
git_ci("_targets.R", "Add new targets to run.")
```

We're now ready to start building our first item/product! But first,
since we finished writing some code, let's style `_targets.R` using
{{< var keybind.styler >}}. Before continuing, let's commit the changes
we've made (including the files in the `_targets/` folder) by putting it
to the Git history with {{< var keybind.git >}}. Then push the changes
to GitHub.

```{r purl-only-commit-packages-to-targets}
#| eval: false
#| echo: false
#| purl: true
git_ci(
  c("_targets", "_targets.R", "DESCRIPTION"),
  "Tell targets about packages, add _targets"
)
```

## :book: Reading task: Fixing issues in the stored pipeline data

**Time: \~10 minutes.**

Sometimes you need to start from the beginning and clean everything up
because there's an issue that you can't seem to fix. In this case,
`{targets}` has a few functions to help out. Here are four that you can
use to delete stuff (also [described on the targets
book](https://books.ropensci.org/targets/data.html#cleaning-up-local-internal-data-files)):

`tar_invalidate()`

:   This removes the metadata on the target in the pipeline, but doesn't
    remove the object itself (which `tar_delete()` does). This will tell
    `{targets}` that the target is out of date, since it has been
    removed, even though the data object itself isn't present. You can
    use this like you would `select()`, by naming the objects directly
    or using the `{tidyselect}` helpers. For example,
    `tar_invalidate(everything())` will remove all metadata details for
    all targets in the pipeline, and
    `tar_invalidate(starts_with("data_"))` will remove all metadata
    details for targets that start with `data_` (e.g. like our
    `data_file` target).

`tar_delete()`

:   This deletes the stored objects (e.g. the `lipidomics` or
    `df_stats_by_metabolite`) inside `_targets/`, but does not delete
    the record in the pipeline. So `{targets}` will see that the
    pipeline doesn't need to be rebuilt. This is useful if you want to
    remove some data because it takes up a lot of space, or, in the case
    of GDPR and privacy rules, you don't want to store any sensitive
    personal health data in your project. Use it like
    `tar_invalidate()`, with functions like `everything()` or
    `starts_with()`. For example, `tar_delete(everything())` will delete
    all stored objects in `_targets/` and
    `tar_delete(starts_with("data_"))` will delete all stored objects
    that start with `data_`.

`tar_prune()`

:   This function is useful to help clean up left over or unused objects
    in the `_targets/` folder. You will probably not use this function
    too often.

`tar_destroy()`

:   The most destructive, and probably more commonly used, function.
    This will delete the entire `_targets/` folder for those times when
    you want to start over and re-run the entire pipeline again.

{{< text_snippet sticky_up >}}

## Summary

-   Use a function-oriented workflow together with `{targets}` to build
    your data analysis pipeline and track your pipeline "targets".
-   List individual "pipeline targets" using `tar_target()` within the
    `_targets.R` file and run them with `targets::tar_make()`.
-   Visualize target items in your pipeline with
    `targets::tar_visnetwork()` or list outdated items with
    `targets::tar_outdated()`.
-   Use `targets::tar_read()` to access saved pipeline outputs.

{{< include /includes/_code-appendix.qmd >}}
