# Using the pipeline to build research output {#sec-pipelines-outputs}

```{r setup}
#| include: false
source(here::here("R/functions.R"))
library(tidyverse)
library(targets)
lipidomics <- read_csv(here::here("data/lipidomics.csv"))
```

{{< text_snippet review_note >}}

While creating

## Learning objectives

{{< include /includes/objectives/_pipelines-output.qmd >}}

## Adding our Quarto report as a target in our pipeline

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Visually walk through this section, showing the products we want,
highlighting the diagram and why we make these diagrams, and emphasizing
the function-oriented workflow. Part of a function-oriented workflow is
making these diagrams that explicitly show the steps (the functions)
that transform one object to another.
:::

If you recall from @fig-pipeline-schematic, we had three items (that we
call products) that we wanted to create from our lipidomics data:

1.  A table of basic descriptive statistics, which we'll call this
    `table_descriptive_stats`.
2.  A plot of the continuous lipid variables, which we'll call this
    `plot_distributions`.
3.  A report (Quarto document). We'll call this `report`.

Let's add these products into the design of our `{targets}` pipeline
that we made in @fig-our-pipeline-design. We'll indicate that our first
two targets have been made in the graph, which now looks like what is
shown in @fig-first-two-done-pipeline.

```{mermaid}
%%| label: fig-first-two-done-pipeline
%%| fig-cap: Targets in our pipeline that we've done in the last session and those we will do in this session.
%%| echo: false
graph TB
    data_file["data_file<br>'data/lipidomics.csv'"]:::done -->|"read_csv()"| lipidomics:::done
    lipidomics -->|"create_table_descriptive_stats()"| table_descriptive_stats:::wip
    lipidomics -->|"create_plot_distributions()"| plot_distributions:::wip
    table_descriptive_stats -->|"tar_read()"| report:::wip
    plot_distributions -->|"tar_read()"| report

    classDef done stroke-width:3px
    classDef wip fill:#ffedd6
```

So in this session, we want to make the `table_descriptive_stats` and
`plot_distributions` targets, then include them in the `report`. In
order for us to slowly build this up, we will start at the end first:
rendering our Quarto document to HTML in our `{targets}` pipeline.

Open up the `docs/learning.qmd` file and create a new code chunk at the
bottom of the file. Name the code chunk `setup` (using the chunk label
`{r setup}`).

We'll be using some `{targets}` functions in this document, so let's add
`library(targets)` to the `setup` code chunk. Since we'll eventually be
adding code as functions into the `R/functions.R` file, we'll `source()`
it right now in the `setup` code chunk and using `here::here()` to give
the correct file path to the `R/functions.R` file. Lastly, in order use
the `lipidomics` data that we added to the `_targets.R` pipeline in the
last session (and that is also stored in the `_targets/` target
"store"), we use `tar_read()` to read it in and assign it to the
`lipidomics` object.

```` {.markdown filename="docs/learning.qmd"}
```{{r setup}}
library(targets)
source(here::here("R/functions.R"))
lipidomics <- tar_read(lipidomics)
```
````

Let's try to render this right now by using {{< var keybind.render >}}.
We will likely get an error saying something about not finding the
`{targets}` "store". This is because when we use a `{targets}` function
like `tar_read()`, it will look for a folder called `_targets/` in the
same folder that the R session is running in. When we render a Quarto
document, the default location for the R session is the folder where the
Quarto file is located, which is `docs/`. We don't have a folder called
`docs/_targets/`. Instead, it is in the root of the project. So we have
to tell `{targets}` where to find the store, which we can do by using
`targets::tar_config_set()`. This function allows us to set various
configuration options for `{targets}`. One of these options is `store`,
which we can use to tell `{targets}` where the `_targets/` store is
located relative to the Quarto file.

So in the `setup` code chunk, we'll add the `tar_config_set()` line
right after `library(targets)`. Since the Quarto file is in the `docs/`
folder, we can use the relative path `"../_targets"` to point to the
store.

```` {.markdown filename="docs/learning.qmd"}
```{{r setup}}
library(targets)
tar_config_set(store = "../_targets")
source(here::here("R/functions.R"))
lipidomics <- tar_read(lipidomics)
```
````

Let's try to render the file now with {{< var keybind.render >}}. It
should work now without any errors! :tada:

Great! Now let's include the Quarto file as a target in our pipeline.
Open up the `_targets.R` file. But, how do we add a Quarto file as a
target? If we try to look through the `{targets}` documentation, we
won't find anything specific for Quarto files. However, there is a
helper package called `{tarchetypes}` that contains functions that make
it easier to add specific types of files as targets, including Quarto
files. You can see at the top of the `_targets.R` file that there is a
commented out line for loading the `{tarchetypes}` package. So let's
uncomment that line so that it looks like this now:

``` {.r filename="_targets.R"}
# Load packages required to define the pipeline:
library(targets)
library(tarchetypes)
```

We don't have the `{tarchetypes}` package installed yet, so let's
install it by going to the Console and using
`use_package("tarchetypes")`.

```{r tarchetypes-deps}
#| purl: true
#| eval: false
#| filename: Console
use_package("tarchetypes")
```

Now, let's go back to the `_targets.R` file and go to the bottom of the
file with the `list()` of `tar_target()` items. We will now add the new
item for rendering the Quarto file that is called `tar_quarto()`. Like
`tar_target()`, this takes a `name` argument. But unlike having a
`command` argument like in `tar_target()`, `tar_quarto()` takes a `path`
argument where you write the file path to the Quarto file. Let's add it:

``` {.r filename="_targets.R"}
list(
  ...,
  tar_quarto(
    name = quarto_doc,
    path = "docs/learning.qmd"
  )
)
```

Since we've made some changes, let's first style the file using
{{< var keybind.styler >}}. Now let's run `targets::tar_make()` with
{{< var keybind.targets-make >}} to build the Quarto document. However,
we will probably get an error because we haven't installed the
`{quarto}` R package yet.

<!-- TODO: Confirm error message -->

So let's install the `{quarto}` R package to help `{targets}` connect to
the Quarto file:

```{r quarto-deps}
#| purl: true
#| eval: false
#| filename: Console
use_package("quarto")
```

Now, let's build the pipeline with {{< var keybind.targets-make >}}
again. It should work! :tada: Let's also check the visualization of the
pipeline with {{< var keybind.targets-visual >}}. You should now see the
`quarto_doc` target at the end of the pipeline.

When we run the pipeline with the Quarto target in it, we'll see that a
new file is created in the `docs/` folder. When we use
`targets::tar_config_set(store = ...)`, it creates a new file called
`_targets.yaml` that contains details that tell `{targets}` where to
find the store. Since the path listed in this `docs/_targets.yaml` file
is a relative path to the store, we can add it to our Git history so
that our collaborators (or future you) also use it.

So let's commit the changes we've made to our project into the Git
history with {{< var keybind.git >}}.

```{r purl-only-setup-code-chunk}
#| eval: false
#| echo: false
#| purl: true
basic_stats_md_text <- c(
  "```{r setup}",
  "library(targets)",
  "tar_config_set(store = '../_targets')",
  "source(here::here('R/functions.R'))",
  "lipidomics <- tar_read(lipidomics)",
  "```",
  ""
)
print_file("docs/learning.qmd")
revise_by_line_num(
  "docs/learning.qmd",
  basic_stats_md_text,
  # Update numbers
  remove_original_lines = -30,
  insert_at_line = 29
)
git_ci("docs/learning.qmd", "Add code for basic stats to report.")
```

## Building our first "product": A table of basic descriptive statistics

Time to make our first "product"! Recalling from
@fig-first-two-done-pipeline, we decided to call our product
`table_descriptive_stats` and to call the function to make it
`create_table_descriptive_stats()`. This function will take the
`lipidomics` data as input (the first argument) and return the
`table_descriptive_stats` as output.

To wrangle the `lipidomics` into `table_descriptive_stats`, we'll use
functions from `{tidyverse}`, specifically `{dplyr}`. We don't yet have
`{tidyverse}` added to our dependencies, so let's add it! Because
`{tidyverse}` is a "meta"-package (special type of package that makes it
easy to install and load other packages), we need to add it to the
`"depends"` section of the `DESCRIPTION` file.

```{r add-tidyverse-deps}
#| filename: Console
#| purl: true
#| eval: false
use_package("tidyverse", "depends")
```

::: callout-note
If you try to run this code without using `"depends"`, you'll get an
error that explains why you can't do that.
:::

Next, let's open up the `docs/learning.qmd` file and add
`library(tidyverse)` to the `setup` code chunk, so that we can use the
functions from `{tidyverse}`. Put it at the top of the `setup` code
chunk, like below:

```` {.markdown filename="docs/learning.qmd"}
```{{r setup}}
library(tidyverse)
library(targets)
tar_config_set(store = "../_targets")
source(here::here("R/functions.R"))
lipidomics <- tar_read(lipidomics)
```
````

Then, below this code chunk, we'll create a new header called
`## Making a table` as well as a new code chunk below that with
{{< var keybind.chunk >}}. There's a lot of different descriptive
statistics we could use for the table, as well as different variables to
include in it. But commonly used descriptive statistics are to show the
means and standard deviations (SD) for our variables. In our case, we
could calculate the mean and SD for each metabolite and display those
stats in our table. To make the table more readable, we could also round
the numbers from the statistics to one digit.

We can achieve this with code by using the split-apply-combine
methodology and the `across()` function, which we covered in the
[split-apply-combine](https://r-cubed-intermediate.rostools.org/sessions/split-apply-combine)
session of the intermediate workshop. Representing this visually, where
each node is a data frame, it might look like:

```{mermaid}
flowchart TB
    lipidomics --> by_metabolite --> mean_sd --> rounded_output
```

The functions to get to each object might be something like:

1.  `group_by()` to split the dataset by the metabolites.
2.  `summarise()` to calculate the `mean()` and standard deviation
    (`sd()`) for each metabolite.
3.  `mutate()` to `round()` the calculated mean and standard deviation
    to 1 digit.

We could use these functions and modify or create columns by directly
referring to them. Or we could make the function more robust by using
the `across()` functional so that it applies the same transformations on
multiple columns based on some condition. By using this function, our
code doesn't depend on the exact names of the columns in the data frame,
so if they change names, our code will still work.

If we look at the help documentation of `?across` (and by recalling from
the
[intermediate](https://r-cubed-intermediate.rostools.org/sessions/split-apply-combine)
workshop), we can give `across()` a `list()` that contains multiple
functions to apply to our columns.

::: {.callout-note collapse="true"}
## :teacher: Teacher note

Slowly build up this pipe of functions, running it each time you add a
function to the pipe. Explain what the code is doing and why at each
step.
:::

Let's write out the code we planned in the code chunk we had just
created:

```{r mean-sd-by-each-metabolite}
#| filename: "docs/learning.qmd"
lipidomics |>
  group_by(metabolite) |>
  summarise(across(value, list(mean = mean, sd = sd))) |>
  mutate(across(where(is.numeric), ~ round(.x, digits = 1)))
```

After that, style the file using {{< var keybind.styler >}}. Then,
commit the changes to the Git history with {{< var keybind.git >}}.

We've made this code, but now we want to actually follow a
"function-oriented" approach so that we can make use of it in our
`{targets}` pipeline. Time to try it on your own!

## :technologist: Exercise: Convert descriptive statistics code into a function

**Time: \~20 minutes.**

In the `docs/learning.qmd` file, use the function-oriented workflow, as
taught in the [intermediate
workshop](https://r-cubed-intermediate.rostools.org/sessions/functions#the-basics-of-a-function),
to take the code we wrote above and convert it into a function.

``` r
create_table_descriptive_stats <- function(___) {
  ___
}

# It should output the same thing as before, but using it like:
create_table_descriptive_stats(lipidomics)
```

Complete these tasks:

1.  Convert the code into a function by wrapping it with
    `function() {...}` and name the new function `descriptive_stats`.
    Here is some scaffolding to help you get started:

    ``` {.r filename="docs/learning.qmd"}
    create_table_descriptive_stats <- function(___) {
      ___
    }
    ```

2.  Replace `lipidomics` in the code we wrote before with `data` and put
    `data` as an argument inside the brackets of `function()`.

3.  At the start of each function we use inside our `descriptive_stats`
    function, add the name of the package like so: `packagename::`.
    E.g., add `dplyr::` before `group_by()` because it's from
    the`{dplyr}` package. Remember that you can look up which package a
    function comes from by writing `?functionname` in the Console.

4.  Style the code using {{< var keybind.styler >}} to make sure it is
    formatted correctly. You might need to manually force a styling if
    lines are too long.

5.  With the *cursor* inside the function, add some roxygen
    documentation with {{< var keybind.palette >}} followed by typing
    "roxygen comment". Remove the lines that contain `@examples` and
    `@export`, then fill in the other details (like the `@params` and
    `Title`). In the `@return` section, write "A data.frame/tibble."

6.  Cut and paste the function over into the `R/functions.R` file.

7.  Source the `R/functions.R` file with {{< var keybind.source >}}, and
    then test the code by running `descriptive_stats(lipidomics)` in the
    Console. If it works, do the last task.

8.  Save both files (`learning.qmd` and `functions.R`). Then open the
    Git interface and commit the changes you made to them with
    {{< var keybind.git >}}.

::: {.callout-tip appearance="default" title="Tip: Implicit returns"}
In the intermediate workshop, we suggested using `return()` at the end
of the function. Technically, we don't need an explicit `return()`,
since the output of the last code that R runs within the function will
be the output of the function. This is called an *implicit return* and
we will be using this feature throughout the rest of this workshop.
:::

```{r solution-descriptive-stats}
#| code-fold: true
#| code-summary: "**Click for a potential solution**. Only click if you are struggling or are out of time."
#' Calculate descriptive statistics of each metabolite.
#'
#' @param data The lipidomics dataset.
#'
#' @return A data.frame/tibble.
#'
create_table_descriptive_stats <- function(data) {
  data |>
    dplyr::group_by(metabolite) |>
    dplyr::summarise(dplyr::across(value, list(mean = mean, sd = sd))) |>
    dplyr::mutate(dplyr::across(tidyselect::where(is.numeric), ~ round(.x, digits = 1)))
}

create_table_descriptive_stats(lipidomics)
```

{{< text_snippet sticky_up >}}

## Adding the first "product" function to the pipeline

Now that we've created a function to calculate some basic statistics, we
can now add it as a step in the `{targets}` pipeline. Open up the
`_targets.R` file and go to the end of the file, where the `list()` and
`tar_target()` code are found. In the first `tar_target()`, replace the
target to load the lipidomic data. In the second, replace it with the
`descriptive_stats()` function. If we want to make it easier to remember
what the target output is, we can add `df_` to remind us that it is a
data frame. It should look like:

```{r targets-basic-stats}
#| eval: false
#| filename: "targets.R"
list(
  ...,
  tar_target(
    name = df_stats_by_metabolite,
    command = descriptive_stats(lipidomics)
  )
)
```

## Creating figure outputs

Not only can we create data frames with targets (like above), but also
figures. Let's write some code to create the plot we listed as our
"output 2" in @fig-pipeline-schematic. Since we're using `{ggplot2}` to
write this code, let's add it to our `DESCRIPTION` file.

```{r add-ggplot2-deps}
#| eval: false
#| purl: true
#| filename: Console
use_package("ggplot2")
```

Next, we'll switch back to `docs/learning.qmd` and write the code to
this plot of the distribution of each metabolite. We'll use
`geom_histogram()`, nothing too fancy. And since the data is already in
long format, we can easily use `facet_wrap()` to create a plot for each
metabolite. We use `scales = "free"` because each metabolite doesn't
have the same range of values (some are small, others are quite large).

```{r histogram-metabolites}
#| fig-cap: "Histograms showing the distribution of all metabolites in the lipidomics dataset."
#| filename: "docs/learning.qmd"
metabolite_distribution_plot <- ggplot(lipidomics, aes(x = value)) +
  geom_histogram() +
  facet_wrap(vars(metabolite), scales = "free")
metabolite_distribution_plot
```

We now have the basic code to convert over into functions.

## :technologist: Exercise: Convert the plot code to a function

**Time: \~10 minutes.**

For now, we will only take the code to make the distribution plot and
convert it into a function. Just like you did with the
`descriptive_stats()` function in the exercise above, complete these
tasks:

1.  Wrap the plot code inside `docs/learning.qmd` with
    `function() {...}` and name the new function `plot_distributions`.
    Use this scaffolding code to help guide you to write the code into a
    function.

    ``` {.r filename="docs/learning.qmd"}
    plot_distributions <- function(___) {
      ___
    }
    ```

2.  Replace `lipidomics` with `data` and put `data` as an argument
    inside the brackets of `function()`.

3.  Add `ggplot2::` to the start of each `{ggplot2}` function used
    inside your function.

4.  Style using {{< var keybind.styler >}} to make sure it is formatted
    correctly. You might need to manually force a styling if lines are
    too long.

5.  With the *cursor* inside the function, add some roxygen
    documentation with {{< var keybind.roxygen >}}. Remove the lines
    that contain `@examples` and `@export`, then fill in the other
    details (like the `@params` and `Title`). In the `@return` section,
    write "A plot object."

6.  Cut and paste the function over into the `R/functions.R` file.

7.  Source the `R/functions.R` file ({{< var keybind.source >}}) and
    then test the code by running `plot_distributions(lipidomics)` in
    the Console. If it works, do the last task.

8.  Save both files and then open the Git interface and commit the
    changes you made to them with {{< var keybind.git >}}.

```{r solution-new-function-descriptive-plots}
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
## This should be in the R/functions.R file.
#' Plot for basic distribution of metabolite data.
#'
#' @param data The lipidomics dataset.
#'
#' @return A ggplot2 graph.
#'
plot_distributions <- function(data) {
  data |>
    ggplot2::ggplot(ggplot2::aes(x = value)) +
    ggplot2::geom_histogram() +
    ggplot2::facet_wrap(ggplot2::vars(metabolite), scales = "free")
}
```

::: {.callout-note appearance="minimal" collapse="true" icon="false"}
## :technologist: Extra exercise: Test out how `tar_outdated()` and `tar_visnetwork()` work

**Time: \~10 minutes.**

Let's make a change to our function and test out how the
`tar_outdated()` and `tar_visnetwork()` work.

1.  Open up the `R/functions.R` file and go to the `descriptive_stats()`
    function.
2.  Add median and interquartile range (IQR) to the `summarise()`
    function, by adding it to the end of `list(mean = mean, sd = sd)`,
    after the second `sd`. Note, IQR should look like `iqr = IQR` since
    we want the output columns to have a lowercase for the column names.
3.  Run `tar_outdated()` and `tar_visnetwork()` in the Console (or by
    using the Command Palette {{< var keybind.palette >}}, then "targets
    outdated" or "targets visual"). What does it show?
4.  Style using {{< var keybind.styler >}}.
5.  Run `tar_make()` in the Console or with
    {{< var keybind.targets-make >}}. Re-check for outdated targets and
    visualize the network again.
6.  Open up the Git interface and commit the changes to the Git history
    with {{< var keybind.git >}}.

```{r solution-new-function-descriptive-stats}
#| code-fold: true
#| code-summary: "**Click for a potential solution**. Only click if you are struggling or are out of time."
#' Calculate descriptive statistics of each metabolite.
#'
#' @param data Lipidomics dataset.
#'
#' @return A data.frame/tibble.
#'
descriptive_stats <- function(data) {
  data |>
    dplyr::group_by(metabolite) |>
    dplyr::summarise(dplyr::across(value, list(
      mean = mean,
      sd = sd,
      median = median,
      iqr = IQR
    ))) |>
    dplyr::mutate(dplyr::across(tidyselect::where(is.numeric), ~ round(.x, digits = 1)))
}
```
:::

{{< text_snippet sticky_up >}}

## Adding the plot function as pipeline targets

Now, let's add the plot function to the `_targets.R` file. Let's write
this `tar_target()` item within the `list()` inside `_targets.R`. To
make it easier to track things, add `fig_` to the start of the `name`
given.

``` {.r filename="targets.R"}
list(
  ...,
  tar_target(
    name = fig_metabolite_distribution,
    command = plot_distributions(lipidomics)
  )
)
```

First, style the file using {{< var keybind.styler >}}. Next, test that
it works by running `targets::tar_visnetwork()` using
{{< var keybind.targets-vis >}} or running `targets::tar_outdated()`
with {{< var keybind.targets-outdated >}}. You should see that the new
item is "outdated". Then run `targets::tar_make()` using
{{< var keybind.targets-make >}} to update the pipeline. If it all
works, than **commit the changes to the Git history** with
{{< var keybind.git >}}.

```{r purl-only-plot-fns-in-targets}
#| eval: false
#| echo: false
#| purl: true
update_targets_plots <- "
),
tar_target(
  name = fig_metabolite_distribution,
  command = plot_distributions(lipidomics)
)
"
# print_lines("_targets.R")
# -20 to remove the previous `)`
revise_by_line_num(
  path = "_targets.R",
  insert_text = update_targets_plots,
  remove_original_lines = -42:-43,
  insert_at_line = 41
)

styler::style_dir()
targets::tar_visnetwork()
targets::tar_make()
git_ci("DESCRIPTION", "Add ggplot2 to deps")
git_ci("R/functions.R", "Add output switch function, to use for targets")
git_ci("_targets.R", "Update targets with plot item")
```

For the `df_stats_by_metabolite`, we can do some minor wrangling with
`mutate()` and `glue::glue()`, and than pipe it to `knitr::kable()` to
create a table in the output document. The `{glue}` package is really
handy for formatting text based on columns. If you use `{}` inside a
quoted string, you can use columns from a data frame, like `value_mean`.
So we can use it to format the final table text to be
`mean value (SD value)`:

```{r stats-to-table}
#| eval: false
#| filename: "docs/learning.qmd"
targets::tar_read(df_stats_by_metabolite) |>
  mutate(MeanSD = glue::glue("{value_mean} ({value_sd})")) |>
  select(Metabolite = metabolite, `Mean SD` = MeanSD) |>
  knitr::kable(caption = "Descriptive statistics of the metabolites.")
```

```{r purl-only-stats-to-table}
#| eval: false
#| echo: false
#| purl: true
pretty_basic_stats_code <- '
targets::tar_read(df_stats_by_metabolite) |>
  mutate(MeanSD = glue::glue("{value_mean} ({value_sd})")) |>
  select(Metabolite = metabolite, `Mean SD` = MeanSD) |>
  knitr::kable(caption = "Descriptive statistics of the metabolites.")
'
revise_by_text(
  path = "docs/learning.qmd",
  original = "tar_read\\(df_stats_by_metabolite\\)",
  replacement = pretty_basic_stats_code
)
styler::style_file("docs/learning.qmd")
git_ci("docs/learning.qmd", "Basic stats as a pretty table.")
```

```{r}
#| echo: false
lipidomics |>
  descriptive_stats() |>
  mutate(MeanSD = glue::glue("{value_mean} ({value_sd})")) |>
  select(Metabolite = metabolite, `Mean (SD)` = MeanSD) |>
  knitr::kable(caption = "The mean and standard deviation of metabolites in the lipidomics dataset.")
```

Rerun `targets::tar_visnetwork()` using {{< var keybind.targets-vis >}}
to see that it now detects the connections between the pipeline targets.
Then, run `targets::tar_make()` with {{< var keybind.targets-make >}}
again to see everything re-build! Last things are to re-style using
{{< var keybind.styler >}}, then **commit** the changes to the Git
history before moving on with {{< var keybind.git >}}. Then push your
changes up to GitHub.

## Summary

-   Within R Markdown / Quarto files, use `targets::tar_read()` to
    access saved pipeline outputs. To include the Quarto in the
    pipeline, use `{tarchetypes}` and the function `tar_quarto()`.

{{< include /includes/_code-appendix.qmd >}}
