# Smoother project-based collaboration {#sec-smoother-collaboration}

```{r setup}
#| include: false
source(here::here("R/functions.R"))
```

Many of you probably work mostly on your own, but as you move through
your career (whether in academia or industry), you will need to---and
perhaps want to---*directly collaborate*[^smoother-collaboration-1] much
more with others.

[^smoother-collaboration-1]: Here, "directly collaborate" means
    contributing to a shared project directly (like editing the same
    files), rather than discussion- or planning-based collaborations
    (and definitely not the "emailing-files-around" type of
    collaboration).

Even though different types of collaboration (e.g. meetings,
brainstorming, real-time co-writing) form the basis for almost all
research-based work (and probably most non-research-based work), direct
collaboration quickly becomes unmanageable when relying on traditional
academic "workflows" (like emailing files around). That's when you'll
need tools designed for collaboration, such as Git and GitHub. But
that's just the starting point; many other factors related to workflows
and processes must be considered to collaborate more effectively.

This session focuses on using more automated ways to structure data
analysis projects and facilitate smoother collaboration.

## Learning objectives {#sec-smoother-collaboration-learning-objectives}

This session’s **overall learning outcome** is to:

1.  Identify actions to streamline collaboration on a data analysis
    project and create projects with R that apply many of these actions.

**Specific objectives** are to:

1.  Explain what a "project-oriented" workflow is, what a project-level
    R dependency management is, and why these concepts are important for
    collaborative and reproducible analyses.
2.  Describe the difference between workflow dependencies and build
    dependencies, and apply functions in the `{usethis}` R package to
    implement these dependency management concepts.
3.  Explain how following a style guide helps build a common approach to
    reading and writing R code, and thereby improves project-level
    collaboration.
4.  Use `{styler}` and RStudio's canonical markdown mode to
    programmatically check and apply style guides to your project files.

## :speech_balloon: Discussion activity: How do you *exactly* collaborate or contribute to your own or others' projects?

> Time: \~10 minutes.

Reflect on when you work on a project (for your thesis or a
manuscript), how **exactly** do you and your collaborators contribute to
the project:

-   Is it mostly verbal contributions?
-   Do you use a shared folder that the files are on?
-   How do you keep track of who's changed what?
-   Do you mostly work on your own with contributions being mostly
    verbal or written feedback (like in a meeting or through an email)?
-   If you collaborate directly on a project, how do you coordinate
    things? Does one collaborator work on one section or analysis, so
    your files are separate?
-   Do you ever have to go in and contribute your own code to theirs
    (and vice versa)?

Consider these questions as we do the following steps.
1.  Take about 1 minute to reflect on these questions.
2.  For 6 minutes, discuss these questions with your neighbour, and talk
    about your own experiences.
3.  For the remaining time, we will briefly share with everyone.

## Project-level package dependency management

> Note: This first session is more conceptual and is heavier on the
> reading and explanation, but is important for the next sessions.

::: callout-note
## :book: Reading task: \~8 minutes

One of the first things to consider when working collaboratively on a
data analysis project (and probably other types of projects too) is what
software to use for your project. This starts out at the highest level:
Are you using R or some other software for the analysis? Since this is
an R course, we're assuming the software will be R! 😜

The next consideration is determining which packages your project will
depend on to produce the results. When working collaboratively with
others (or even with your future self), you'll need a clear way to track
the project's package dependencies, and, ideally, have a simple
and efficient method to install or update them.

Part of the approach covered in this course requires that you follow a
["project-oriented"
workflow](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/)
when working on, well, a project. In order to know how to track your
project's package dependencies, you need to first know, what a "project"
is and how do we work around it. Since the introduction course's first
session on the [Management of R
Projects](https://r-cubed-intro.rostools.org/sessions/project-management.html),
we've consistently taught and used this workflow-style. In fact, it is
embedded into the use of the R Projects via the `.Rproj` files and in
the use of the `{here}` package. So, we're already following a approach,
which will make it easier to track package dependencies of our project.

![Cartoon of combining R Projects with the `{here}` package, compared to
the common approach of using `setwd()`. [Artwork by
\@allison_horst.](https://allisonhorst.com/allison-horst)](https://cdn.myportfolio.com/45214904-6a61-4e23-98d6-b140f8654a40/e684ee45-98fc-469e-997c-e2d68b0096cd_rw_1920.png?h=d754f634e6fc1e292430124fbe44dad7){fig-alt="The R Project text is made of metal,
and the `{here}` package wears a helmet, appearing secure and robust and
ready to go skateboarding. Meanwhile `setwd()` is made of glass, covered
in casts and band-aids, looking much more fragile."}

Let's start with the `AdvancedR3` project, created in @sec-pre-course,
that includes the `lipidomics` dataset. We have code in the
`data-raw/nmr-omics.R` file already that uses some packages. Let's
assume that your project will be more complex than this and that you
will eventually need some collaborators to contribute who are experts in
for instance metabolomics data processing and in statistical analysis of
high-dimensional data. You know you will end up needing to use other
packages. You also know that you all need some way of tracking which
packages are used so that when others join and contribute to the
project, they can as seamlessly as possible install or update the
packages your data analysis project needs. There are a few ways of
"tracking" package dependencies.

The simplest, but most primitive way is to always make sure to use
`library()` at the top of each R script for each package that the R
script uses. We can call this "*informal*" dependency management.

Let's review the advantages and disadvantages of this form of dependency
management:

-   Advantage:

    -   It's the easiest to conceptually understand and to use.

-   Disadvantages:

    -   It doesn't track project-level dependencies very well, since
        multiple scripts might use the same packages, while some might
        use different ones. This means that you can't easily and quickly
        install or update all the packages used in your project. You
        will probably have to go through each R script manually and
        install each package manually.

    -   It doesn't track the *versions* of the packages your project
        depends on, so if a package gets updated and it breaks
        something, you might not be able to figure out how to quickly
        fix that issue, especially for those deadline crunches.

You might have seen a similar "informal" approach where scripts starts
with code that looks like this:

``` r
if (!require("packagename")) {
  install.packages("packagename")
}
```

This code checks whether a package exists (i.e., is installed), and if
not, it installs it. But! This is not an optimal method to track
packages because `require()` won't load the package if it doesn't find
it. As a result, you will have to rerun the script a few times if you
don't have that package installed; first, to install the package, then
to load it. Plus, sometimes you may need to restart the R session after
installing a package in order for R to detect it properly. So, this is
not a very efficient way to track dependencies either.

What's the alternative, then? One alternative is a more *formal* way of
managing your dependencies, which is what we'll cover in the next
section.

{{< include ../includes/_sticky_up.qmd >}}
:::

## Formal dependency management

<!-- TODO: As a reading activity or explained by the instructor? -->

:::: callout-note
## :book: Reading task: \~3 minutes

The most common form of a more *formal* way of managing dependencies---
at least based on R packages and projects found on GitHub---makes use of
the `DESCRIPTION` file and the `usethis::use_package()`. We covered this
style of dependency in the [intermediate
course](https://r-cubed-intermediate.rostools.org/dry-functions.html#continuing-the-workflow).
We will also use this approach during this course, but we'll expand on
it a lot more. We use and recommend it for managing dependencies because
many tools and workflows rely on it, making it well-integrated with R
projects.

Let's take a look on the advantages and disadvantages of this way of
tracking dependencies:

-   Advantages:

    -   Relatively easy to conceptually understand, since you can
        directly view the packages used in the by opening the
        `DESCRIPTION` file and looking at the contents.

    -   Because it is widely used, there are many processes already
        built around making use of tracking dependencies this way. For
        instance, when creating an R package, you need to track package
        dependencies this way.

    -   Installing packages is as easy as opening the project and
        running `pak::pak()` in the Console, which will install all the
        packages listed in the `DESCRIPTION` file.

    -   Adding packages is as easy as writing
        `usethis::use_package("packagename")` in the Console.

-   Disadvantages:

    -   Like the previous method, it doesn't automatically keep track of
        the versions of the packages you are using. Even though it is
        possible to add a specific version to the `DESCRIPTION` file,
        `{pak}` will install the newest version of the package.

    -   Your project might still rely on a package that is installed on
        *your* computer which influences your project. It could be that
        it isn't obvious as a dependency, so you forget to include it.
        Meaning, you are still responsible for adding the project
        dependencies to the `DESCRIPTION` file.

::: {.callout-tip appearance="default"}
There is an even more formal and structured way of tracking dependencies
using the [renv](https://rstudio.github.io/renv/index.html) package,
that we covered in previous versions of this course. It is a great
package to use, especially if you are working on a project with several
others or a project that requires tracking of not just the packages but
their specific versions. We don't teach it anymore because it is quite
complex and difficult to understand. Moreover, for the majority of
projects, it's not strictly necessary.
:::

As we work on the project during the course and realize we need to use a
specific package, we will continue using `usethis::use_package()` to
install it and add it to the `DESCRIPTION` file.

{{< include ../includes/_sticky_up.qmd >}}
::::

Before moving on to the exercise, be sure to add and commit all the
files in the `AdvancedR3` project to save the current state of the files
to the Git history. Open the Git interface by either typing with
{{< var keybind.git >}} or by going to the Git pane and clicking the
"Commit" button. Then, write a commit message and click the "Commit"
button in this window.

## :technologist: Exercise: Add packages from the data processing script {#sec-data-processing-dependency-exercise}

> Time: \~10 minutes.

Let's update the `DESCRIPTION` file with the packages we depend on in
the `data-raw/nmr-omics.R` script. Open `nmr-omics.R` file and complete
the following tasks:

1.  **Manually** look for package dependencies in the R script that are
    declared with `library()` or `::`. It can help to use the "Find in
    files" feature in RStudio to look for all places that have either
    `::` or `library`. Use either `Edit -> Find in Files ...`, or with
    {{< var keybind.find >}} to open up the search popup.
2.  Use `?usethis::use_package` to review how to use this function.
3.  In the Console, run `usethis::use_package(packagename)` for each
    package you find in `data-raw/nmr-omics.R` (from 1. above).
4.  Once done, open the Git interface with {{< var keybind.git >}} (or
    going to the Git Pane and clicking the "Commit" button). What has
    been changed? Commit those changes to the Git history with a
    descriptive message.

```{r solution-add-processing-packages}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are really struggling or are out of time for the exercise."
#| purl: true
#| filename: Console
usethis::use_package("stringr")
usethis::use_package("readxl")
usethis::use_package("dplyr")
usethis::use_package("tidyr")
usethis::use_package("snakecase")
usethis::use_package("here")
usethis::use_package("fs")
usethis::use_package("readr")
```

::: {.callout-note collapse="true"}
## :teacher: Instructor note

Make sure that everyone has added the right packages, since it can be
easy to miss some of the packages referenced using `::`.
:::

```{r purl-only-commit-desc}
#| eval: false
#| echo: false
#| purl: true
usethis::use_git_ignore(".Rbuildignore")
git_ci(
  c("DESCRIPTION", ".gitignore"),
  "Add packages as dependencies to project."
)
```

## Two types of dependencies: Build and workflow

::: {.callout-note appearance="minimal" collapse="true"}
## :teacher: Instructor note

Verbally explain the content of the next few paragraphs (don't need to
show the website on the projector).
:::

When you work on a research project that involves data analysis, you'll
likely use packages in two different ways:

1.  Packages that *directly* contribute to data wrangling, analysis,
    plotting, and making the manuscript. These types of packages are
    generally called "build" or "deploy" dependencies. Packages like
    `{dplyr}` or `{tidyr}` are usually **build** dependencies, since you
    use them for processing data.
2.  Packages that *assist* you in doing your work but aren't directly
    used for data analysis. These types of packages would be called
    "workflow" or "development" dependencies. `{usethis}` is usually
    considered a **workflow** dependency.

::: {.callout-tip appearance="default" title="Is this package a build or workflow dependency?"}
A good rule of thumb to determine whether a package is a build or a
workflow dependency is:

-   If you write and use functions from the package within an R script
    that does something to the data or analysis, then it is likely a
    *build* dependency.
-   If you *only* ever use functions from the package in the Console,
    then it is probably a *workflow* dependency.
:::

So, now we know that the dependencies we added in the exercise are
actually build dependencies, since we use them directly in our project
to process the data.

The way you add packages to the `DESCRIPTION` file is slightly different
depending on what type of dependency it is. For build dependencies, we
use the function we've already used before:
`usethis::use_package("packagename")`. For workflow dependencies, it's
the same function with a small twist. Our first workflow dependency is
`{usethis}`, which we will use many times throughout the course.

But before we get into adding `{usethis}` to the `DESCRIPTION` file,
let's tweak our personal (also called the global) `.Rprofile` to make our lives a bit easier. We'll add a
setting so we don't have to type `usethis::` before each function.

Run the `usethis::use_usethis()` function to open your own (user)
`.Rprofile`:

```{r add-usethis-rprofile}
#| eval: false
#| filename: Console
usethis::use_usethis()
```

Then, copy and paste the code below into your `.Rprofile`, at the bottom
of the file.

``` r
if (interactive()) {
  suppressMessages(require(usethis))
}
```

Let's restart R with {{< var keybind.restart-r >}}. Now, we're ready to
use `use_package()` to add `{usethis}` as a workflow dependency. The
only difference between adding a build and a workflow dependency is that
for workflow dependencies, we need to add a second argument to
`use_package()` to specify that it is a workflow dependency. The second
argument is `"suggests"`, like so:

```{r suggests-dep}
#| filename: Console
#| eval: false
#| purl: true
use_package("usethis", "suggests")
```

```{r purl-only-update-desc-with-usethis}
#| eval: false
#| echo: false
#| purl: true
# Admin: Restart R session first.
git_ci("DESCRIPTION", "Add usethis as a suggests dependency.")
```

::: {.callout-tip appearance="default"}
If you open the help file for `use_package()` you'll see that the second
argument we used above is called `type`, and "Imports" is its default
value. So, if you don't specify the second argument, the package will be
added under "Imports" in the `DESCRIPTION` file. This is exactly what we
did in the exercise in @sec-data-processing-dependency-exercise!

So, "imports" corresponds to build dependencies, while "suggests" are
equivalent to what we call workflow dependencies.
:::

Open the Git interface and see that under `Suggests:` in the
`DESCRIPTION` file is `usethis`. Let's commit these changes, with
{{< var keybind.git >}}.

::: {.callout-note collapse="true"}
## :teacher: Instructor note

For the information block below, mention it to the learners but you
don't need to go over it. Especially mention the second part of the tip.
:::

::: {.callout-tip appearance="default"}
When you come back to a project after a few months or if you start
collaborating on a project, you can run `pak::pak()` to install all
types of dependencies, both workflow and build. Neat! :tada: 
:::

## :technologist: Exercise: Connect your project to GitHub

> Time: \~25 minutes.

Since we will eventually connect our project Git repository to GitHub to
practice working with GitHub, we'll connect our project to GitHub right
now.

Remember that GitHub is a platform for sharing your projects in a way
that is transparent (if you add all the relevant files and keep a clean
history of your files) and makes it easy to collaborate with others.
It's a great way to share your project with others, get feedback, and
even contribute to other projects.

Let's complete these tasks to connect to GitHub.

1.  First, commit the latest changes to the Git history with
    {{< var keybind.git >}}.

2.  If you haven't yet, please create a
    [GitHub](https://github.com/join) account.

3.  Add the `{gitcreds}` package as a workflow dependency. You'll need
    it for the next item

    ``` r
    use_package("gitcreds", "suggests")
    ```

4.  Read through and complete the tasks in the [*linking your project to
    GitHub* of the *Connect to GitHub*
    Guide](https://guides.rostools.org/connect-github#linking-your-project-to-github).

5.  Check your GitHub to make sure the project repository has been
    uploaded to it.

## Automatically adhere to a style guide

::: callout-note
## :book: Reading task: \~5 minutes

We've covered how to adhere to a style guide in both the introductory as
well as the intermediate course, because it is such a useful and
powerful tool to easily write more readable code. We'll expand on it in
this course because it fits precisely with the theme of collaboration.
That's because, when you're working on your own and don't need to worry
about anyone seeing your code, there's a natural temptation to write
your code like you might write notes to yourself: scribbled and scrawled
down quickly.

But when working with others, how it looks can greatly impact how
quickly and easily others are able to read and interpret your code.
Multiply all the collaborators on a project with this natural temptation
for quick (and potentially sloppy) coding, you can imagine how easy it
is for code to massively "drift" towards being poorly formatted,
especially when deadlines are approaching.

That's when "linters" or "stylers" (types of static code analysis tools)
become very useful. They scan your code for common mistakes or syntax
issues, either listing them for you to fix or automatically fixing them
for you. Linters are especially useful when collaborating with someone
who is less experienced with coding or who contribute occasionally to
assist in the code being consistently formatted across the project.

To ensure consistent styling without needing to run checks manually,
*automatic* linting/styling can be a great solution. That's where the
`{styler}` package comes in!

{{< include ../includes/_sticky_up.qmd >}}
:::

Since we will use `{styler}` for the project as a workflow dependency, let's 
add it to the `DESCRIPTION` file:

```{r add-styler-dep}
#| eval: false
#| purl: true
#| filename: Console
use_package("styler", "suggests")
```

```{r purl-only-update-desc-with-styler}
#| eval: false
#| echo: false
#| purl: true
git_ci("DESCRIPTION", "Add styler as a suggests dependency.")
```

There are only a few functions in `{styler}` that we can to use. The
first is used to style a single file with R code in it using
`styler::style_file()`. However, an easier way to style the file we're
currently working on is the "style active file" RStudio addin. We use
that through the Command Palette ({{< var keybind.palette >}}) and
typing "style file", which should show the "Style active file" option.
This is what we will do frequently throughout the course.

Let's try it out. While inside the `data-raw/nmr-omics.R` file, use
{{< var keybind.styler >}} to style the file. Notice that no changes
were made since the file is already tidy.

::: {.callout-note appearance="default"}
You will probably be asked to install a package called `{miniUI}`, click
"Yes".
:::

If you wanted to run `{styler}` on all the files, we can use:

```{r style-dir}
#| eval: false
#| filename: Console
styler::style_dir()
```

::: {.callout-note collapse="true"}
## :teacher: Instructor note

Mention the callout block below, but don't go into it at all.
:::

::: {.callout-tip appearance="default"}
You might be used to using 4 spaces for tabs instead of 2. The `{tidyverse}`
style uses 2, so that's default option in `{styler}`. If you want to
change that, you can add the following code to the project's
`.Rprofile`:

```{r styler-options-4-spaces}
options(
  styler.addins_style_transformer = "styler::tidyverse_style(indent_by = 4)"
)
```
:::

Before we move on, let's make sure everything has been committed to the
Git history with {{< var keybind.git >}}.

## Styling Markdown files

For multi-person collaborative projects, having some type of code
styling and checker can really help with standardizing how the code
looks, which ultimately will make it easier to read each other's code
contributions.

But what about for Markdown files? While there isn't a package or
function (yet) that styles the Markdown files, RStudio does have an
option in their Tools to format Markdown into a "canonical form". The
reason for this option is because they've added a "visual editor mode"
to writing R Markdown / Quarto files (which is great if you are more comfortable
with apps like Word). Let's set up the project to automatically reformat
Markdown files.

::: {.callout-warning appearance="default"}
Use this option *only if* you have your project under Git version
control, since it will directly modify and overwrite the contents of the
entire file.
:::

There are two ways of doing it:

1.  Go into
    [`Tools -> Project Options -> R Markdown`](https://rstudio.github.io/visual-markdown-editing/options.html#project-options)
    and change the options "Automatic text wrapping" to "column" (with
    the default 72 width value) and "Write canonical visual mode
    markdown" to "true".
2.  Or set the [YAML
    options](https://rstudio.github.io/visual-markdown-editing/markdown.html#writer-options)
    in either the project-level `_quarto.yml` file or at the file-level
    in the YAML header.

For right now, we will do the project option settings so that as long as
we are using RStudio and in the R Project, it will automatically
reformat the Markdown files as you write in them. Follow the
instructions in item 1 above to set the options.

Now, when you save your file, RStudio should automatically reformat the
Markdown into a standardized format. If you want to switch to using the
Visual Mode, use {{< var keybind.visual-edit >}} or the "Visual" button
at the top of the Source Pane beside the bolding and italicizing
buttons.

The instructors won't be using the Visual Mode during the course,
however, you are welcome to do so. We will be using the "canonical"
Markdown mode in the "source" (default) mode though.

Let's test it out. While in the `doc/learning.qmd` file, go to the
bottom of the file and type out:

``` markdown
## This is poorly formatted
- Definitely should have an empty space above this list.
- This isn't a list, why not?
```

Save the file. What happens? Lists in Markdown need to have an empty
space above them to work properly (except for when they are below a
header, but in all other cases it needs a space above). With the
canonical mode on, we can get feedback right away that it isn't right.
It gets automatically fixed by adding that empty space.

``` markdown
## This is poorly formatted

-   Definitely should have an empty space above this list.
-   This isn't a list, why not?
```

Since this mode is on automatically now, as we work in the
`doc/learning.qmd` file through the sessions, we'll get lots of
experience using it.

## :technologist: Exercise: Update the README file while using canonical Markdown mode

> Time: \~10 minutes.

Open up the `README.md` file and do all the `TODO` items found inside the file as well as inside the `TODO.md` file. Save
often and watch as the Markdown gets reformatted.

After you are done, commit the changes you made to the Git history with
{{< var keybind.git >}}.

Then delete the `TODO.md`, followed by committing these deletions in the
Git history. Click the "Push" button to push the changes to GitHub.

```{r purl-only-editor-options-to-readme}
#| echo: false
#| eval: false
#| purl: true
fs::file_delete("TODO.md")
gert::git_commit("Don't need this file anymore.")
```

## Summary

In this chapter, we have covered how to do the following in a project
for smoother collaboration (and nicer-looking code and documents):

-   Track your project package dependencies with the `DESCRIPTION` file
    and `{usethis}`.
-   Install the dependencies necessary for your project with
    `pak::pak()`.
-   Follow a style guide by using `{styler}`. Combine with the Command
    Palette ({{< var keybind.palette >}}) to quickly style the code you
    are actively working on.
-   Use RStudio's canonical markdown mode to reformat Markdown into a
    standard format.

{{< include /includes/_code-appendix.qmd >}}
