# Smoother project-based collaboration {#sec-smoother-collaboration}

```{r setup}
#| include: false
source(here::here("R/functions.R"))
```

Many of you probably work largely and most consistently on your own, but
as your move through your career (in academia or industry), you will
need to and maybe also want to *directly
collaborate*[^smoother-collaboration-1] a lot more with others.
Different types of collaboration (e.g. meetings, brainstorming,
real-time co-writing) form the basis for almost all research-based work
and probably most non-research-based work.

[^smoother-collaboration-1]: Collaborate here meaning directly
    contributing to a shared project, rather than discussed or planning
    based collaborations (and definitely not emailing-files-around
    collaboration).

More direct collaboration on a project quickly becomes unmanageable when
using traditional academic "workflows" (emailing around). That's when
you need to start using tools designed for collaboration, like Git. But
Git is just the starting point. There are many many other things to
consider for workflows and processes to effectively collaborate with
others. This session is about making use of more automated ways of
structuring data analysis projects to ease collaboration.

## Learning objectives

The overall objective for this session is to:

1.  Identify potential actions to streamline collaboration on a data
    analysis project and create projects that apply many of these
    actions using R.

More specific objectives are to:

1.  Explain what a "project-oriented" workflow is, what a project-level
    R dependency management is, what a "project environment" is, and why
    these concepts are important to consider in collaborative and
    reproducible analyses.
2.  Describe the difference between "workflow dependencies" and "build
    dependencies".
3.  Apply functions in the `{renv}` and `{usethis}` R packages to
    implement these dependency management concepts.
4.  Explain the role that following a style guide has on building a
    common approach to reading (and writing) R code, and thus improve
    project-level collaboration.
5.  Use `{styler}` and RStudio's canonical markdown mode to
    programmatically check and apply style guides to your project files.

## Exercise: How do you *exactly* collaborate or contribute? To your own or others' projects.

> Time: \~10 minutes.

When you work on a project (for your thesis or a manuscript), how
**exactly** do you and your collaborators contribute to the project? Is
it mostly verbal contributions? Do you use a shared folder that the
files are on? How do you keep track of who's changed what? Do you mostly
work on your own and contributions are largely verbal or written
feedback (like in a meeting or through an email)? If you work directly
on a project, how do you coordinate things? Does one collaborator work
on one section or analysis, so your files are separate? Do you ever have
to go in and contribute your own code to theirs (and vice versa)?

1.  Take 2 minutes to think on these questions.
2.  For 6 minutes, discuss these questions with your neighbour, and talk
    about your own experiences.
3.  For the remaining time, we will share briefly with everyone.

## Project-level R dependency management

> Note: This first session is more conceptual and is heavier on the
> reading and explanation, but is important for the next sessions.

::: callout-note
## Reading task: \~8 minutes

One of the first things to consider when working collaboratively on a
data analysis project (and probably other types of projects too) is what
software to use for your project. This starts out at the highest level:
Are you using R or some other software for the analysis? Since this is
an R course, we're assuming the software will be R! 😜

The next consideration is which packages your project depends on to
produce the results. When working collaboratively with others, and
yourself several months in the future, you need some way of knowing how
to easily and quickly install or update these package dependencies.

Part of this approach requires that you follow a ["project-oriented"
workflow](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/)
when working on, well, your project. In order to know how to track your
project's package dependencies, you need to first know, what is a
"project" and how do we work around it? Since the introduction course's
first session on the [Management of R
Projects](https://r-cubed-intro.rostools.org/sessions/project-management.html),
we've consistently taught and used this workflow-style. In fact, it is
embedded into the use of the R Projects via the `.Rproj` files and in
the use of the `{here}` package. So we already are following this
approach from the start, which will make it easier to track package
dependencies of our project.

![Cartoon of combining R Projects with the `{here}` package, compared to
the common approach of using `setwd()`. [Artwork by
\@allison_horst.](https://allisonhorst.com/allison-horst)](https://cdn.myportfolio.com/45214904-6a61-4e23-98d6-b140f8654a40/e684ee45-98fc-469e-997c-e2d68b0096cd_rw_1920.png?h=d754f634e6fc1e292430124fbe44dad7)

Let's start with the `AdvancedR3` project that uses the `lipidomics`. We
have code in the `data-raw/nmr-omics.R` file that uses some packages.
Let's assume that your project will be more complex than this and that
you will eventually need some collaborators to contribute who are
experts in for instance metabolomics data processing and in statistical
analysis of high-dimensional data. You know you will end up needing to
use other packages. You also know that you all need some way of tracking
which packages are used so that when others join and contribute to the
project, they can as seamlessly as possible install or update the
packages your data analysis project needs. There are a few ways of
"tracking" package dependencies.

1.  The simplest, but most primitive way is to always make sure to use
    `library()` at the top of each R script for each package that the R
    script uses.

    -   Advantage:

        -   This is the easiest to conceptually understand and to use.

    -   Disadvantages:

        -   It doesn't track project-level dependencies very well, since
            multiple scripts probably use similar packages across them.
            Which means you can't easily and quickly install or update
            all the packages your project uses, since you will probably
            have to go through each R script manually and install each
            package manually. You might have seen some scripts with code
            that looks like this at the top:

            ``` r
            if (!require("packagename")) {
              install.packages("packagename")
            }
            ```

            This code checks if a package exists, if not, it installs
            it. But! This is not an optimal method to track packages
            because `require()` won't load the package if it doesn't
            find it. Which means you would have to re-run the script
            probably a few times. Plus, sometimes you may need to
            restart the R session after installing a package in order
            for R to detect it properly.

        -   It doesn't track the *versions* of the packages your project
            depends on, so if a package gets updated and it breaks
            something, you might not be able to figure out how to
            quickly fix that issue, especially for those deadline
            crunches.

2.  The most common form, at least based on R packages and projects
    found on GitHub, makes use of the `DESCRIPTION` file and
    `usethis::use_package()` to track if a package is used for a project
    or not. We covered this style of dependency in the [intermediate
    course](https://r-cubed-intermediate.rostools.org/dry-functions.html#continuing-the-workflow).
    We will also use this approach during this course, but expand a lot
    more on it.

    -   Advantages:

        -   Relatively easy to conceptually understand, since you can
            directly view the packages your project needs by opening the
            `DESCRIPTION` file and looking at the contents.

        -   Because it is widely used, there are many processes already
            built around making use of tracking dependencies this way.
            For instance, you need to track package dependencies when
            creating R packages.

        -   Installing packages is as easy as opening the project and
            running `remotes::install_deps()` in the Console, which will
            install all the packages listed in the `DESCRIPTION` file.

        -   Adding packages that you need is as easy as writing
            `usethis::use_package("packagename")` in the Console.

    -   Disadvantages:

        -   Like the previous method, it doesn't easily keep track of
            the versions of the packages you are using.

        -   Your project might still rely on a package that is installed
            on *your* computer and that influences your project, but
            that might not be obvious as a dependency or that you forgot
            to include.
:::

Before continuing to the exercise, we need to make sure to add and
comment all the files from the project into the Git history. Open the
Git interface by either typing with {{< var keybind.git >}} or by going
to the Git pane and clicking the "Commit" button.

## Exercise: Add packages from the data processing script

> Time: \~10 minutes.

Since the `DESCRIPTION` file will be used later on for the more formal
dependency management, let's get it updated with the packages we are
using in the `data-raw/nmr-omics.R` script. Open that file and complete
these tasks:

1.  **Manually** look for package dependencies in the R script that are
    declared with `library()` and `::`. It can help to use the "Find in
    files" feature in RStudio to look for all places that have either
    `::` or `library`. Use either `Edit -> Find in Files ...`, or with
    {{< var keybind.find >}} to open up the search popup.
2.  Use `?usethis::use_package` to review how to use this function.
3.  In the Console, run `usethis::use_package()` for each package you
    find in `data-raw/nmr-omics.R` (from 1. above).
4.  Since we are also using the `{usethis}` package to manage these
    things, run `usethis::use_package("usethis")` to also add it as a
    package dependency.
5.  Once done, open the Git interface with {{< var keybind.git >}} (or
    going to the Git Pane and clicking the "Commit" button). What has
    been changed? Commit those changes to the Git history .

```{r solution-add-processing-packages}
#| eval: false
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are really struggling or are out of time for the exercise."
#| purl: true
usethis::use_package("stringr")
usethis::use_package("readxl")
usethis::use_package("dplyr")
usethis::use_package("tidyr")
usethis::use_package("snakecase")
usethis::use_package("here")
usethis::use_package("fs")
usethis::use_package("readr")
usethis::use_package("usethis")
```

::: {.callout-note collapse="true"}
## Instructor note

Make sure that everyone has added the right packages, since it can be
easy to miss some of the packages referenced using `::`.
:::

```{r purl-only-commit-desc}
#| eval: false
#| echo: false
#| purl: true
usethis::use_git_ignore(".Rbuildignore")
git_ci(
  c("DESCRIPTION", ".gitignore"),
  "Add packages as dependencies to project."
)
```

## Formal dependency management

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Briefly verbally go over this text (you don't need to show the website,
you can remain in RStudio).
:::

While the approach of managing package dependencies through the
`DESCRIPTION` file is quite powerful, it has the major disadvantage of
not keeping track of the *version* of each of your packages. So instead
we are going to use a package dedicated to handling project
dependencies, called `{renv}`.

![Cartoon showing a simplified version of what `{renv}` does for R
Projects, by making them more self-contained. [Artwork by
\@allison_horst.](https://allisonhorst.com/allison-horst)](https://cdn.myportfolio.com/45214904-6a61-4e23-98d6-b140f8654a40/3c1eda6b-be13-4257-9c32-4cf865f171ae_rw_1920.png?h=5a6482956a683b2fc417f965674456de)

`{renv}` is a package that manages package dependencies in a project by,
in simple terms, creating a project-specific R "library". You might
think of `library()` when you hear R library, and you aren't completely
wrong. When you call, for instance, `library("usethis")`, R looks for
the package `{usethis}` in your computer's "library" of R packages. This
library can be found by running this function:

```{r libpaths}
.libPaths()
```

::: {.callout-note appearance="default"}
The default libary paths are most likely different on your computer, and
depends on the computer and R setup.
:::

Those file paths are where all R packages are installed to. If there is
more than one path, R checks the first before continuing to the next.
`{renv}` instead creates a file path for R packages to be installed
within the project. This isn't completely what it does, since there are
a lot of very technical details to what `{renv}` does internally, but
this is the basic concept.

In the end, this makes your project relatively self-contained in its
package dependencies.

-   Advantages:

    -   Installing all packages necessary for your project is as easy as
        running `renv::restore()`. `{renv}` even gives you helpful tips
        and instructions when things go wrong or if something is
        missing.

    -   *Every* package, including the packages that your packages
        depend on, have their version tracked. So if a package gets
        updated on CRAN, it doesn't affect you *until you choose*, using
        `renv::update()`. If a package update breaks your code and you
        have a deadline, you can easily go back to the older versions of
        the packages.

    -   Because your project is now self-contained with its own R
        library, it becomes very obvious (through errors) when you might
        be missing some other dependency because your code wouldn't run
        until you install or fix that package dependency.

    -   From a reproducibility point of view, as long as your project is
        tracked by `{renv}`, it's easier to independently have your data
        analysis be reproduced and verified.

-   Disadvantages:

    -   It takes a fair amount of learning to conceptually understand
        what is going on.

    -   When there are issues that come up, it can be difficult to
        figure them out.

    -   Because *all* packages, including those packages that your
        packages depend on, are installed within your project,
        installation times can sometimes be a bit long.

    -   When you restore a library from a `renv.lock` file that uses old
        package versions, it can sometimes be difficult to re-install
        them.

::: margin-content
It's sometimes very annoying to debug these "virtual environments". But
thankfully you can turn it off with `renv::deactivate()`! It still is
worth it to start considering and accounting for how your dependencies
might influence your project results and collaboration.
:::

Let's start using `{renv}` in our `AdvancedR3` project. In the Console,
type out:

```{r renv-init}
#| eval: false
#| purl: true
renv::init(bare = TRUE)
```

```{r purl-only-add-renv-init}
#| eval: false
#| echo: false
#| purl: true
git_ci(
  c("renv", ".Rprofile"),
  "Add renv machinery."
)
```

When prompted, choose to use the explicit declaration.

The function `renv::init()` initializes the project to begin being
managed by `{renv}`. By using the argument `bare = TRUE` we are telling
`{renv}` to not search for dependencies in the project, since we want to
do that ourselves.

When initilising a `{renv}` instance using `bare = TRUE` we create a new
`.libPath()` location that does not contain any of the packages we
previously defined as project dependencies. You can notice this if you
attempt to type `usethis::`, you'll notice that nothing comes up.
Therefore, before being able to continue our workflow we need to tell
`{renv}` to install the packages we have already defined as
dependencies. This can be done with `renv::install()`:

```{r renv-install}
#| eval: false
#| purl: true
renv::install()
```

After doing this we should have several files in our project folder:

```{r show-created-files}
#| echo: false
fs::dir_tree(regexp = "\\.Rprofile|renv", all = TRUE, recurse = 1)
```

-   `.Rprofile`: This is the file that `{renv}` uses to build up its
    machinery. With an `.Rprofile` within the project folder, we can
    create a "project environment". By having this project environment
    that is (mostly) self-contained, it allows us to be a bit closer to
    having a fully reproducible analysis and it makes it easier to
    collaborate, since we all than share the same project setup.

-   `renv.lock`: (Might not be created yet) This contains all the
    information about the packages your project depends on, including
    where it was installed from (CRAN or GitHub for example), what the
    version number is, and more. This is like a supercharged version of
    the `DESCRIPTION` file.

-   `renv/`: This folder contains several other files that make up the
    machinery of `{renv}`. For instance, there is the `library/` folder
    that contains all the R packages necessary for the project. Then
    there are the `activate.R` script and `settings.dcf` file that both
    work to manage the dependencies, like installing, updating, and
    removing.

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Open up the website to this figure to explain how things are working in
the background when using `{renv}`. After explaining the figure and
concepts, you can switch back to RStudio to continue verbally going over
the rest of the text, along with the code, in this section.
:::

```{mermaid fig-renv-environment}
%%| label: fig-renv-environment
%%| fig-cap: Simple schematic of how renv is a separate environment.
%%| echo: false
%%| eval: true
%%{init:{'flowchart':{'nodeSpacing': 40, 'rankSpacing': 20}}}%%
graph LR
    user["User/"] --- root_docs
    user --- root_desktop
    user --- .Rprofile
    subgraph global[Global R environment]
        root_docs["Documents/"] --- r["R/ (global R library)"]
        root_desktop["Desktop/"] --- renv["AdvancedR3/"]
        root_docs --- 3["other-projects/"]
        .Rprofile
        subgraph renv_env[Project R environment]
            renv --- renv_folder["renv/<br/>(R library)"]
            renv --- proj_prof[.Rprofile]
        end
    end

linkStyle 0,1,2,3,4,5,6,7 stroke-width:1px;

classDef folderEnv fill:transparent,stroke-dasharray:5;
class renv_env,global folderEnv
```

**Before we continue, let's commit the new files to the Git history**
with {{< var keybind.git >}}.

The general workflow for using `{renv}` while working on your project is
described in more detail on the [Introduction to
renv](https://rstudio.github.io/renv/articles/renv.html#workflow)
webpage. However, unlike the general workflow, we also want to continue
using the `DESCRIPTION` file. That's because a lot of tools and
workflows exist that make use of it, so we want to remain compatible
with them.

As we work on the project and realize we need to use a specific package,
we would normally use `install.packages()` and then add `library()` to
the script or R Markdown file. Later on, we'd eventually run
`renv::snapshot()` to update the `renv.lock` file with the new packages
we've installed. `renv::snapshot()` as well as `renv::init()` usually
rely on "implicit" dependencies, meaning `{renv}` will search throughout
the project for any packages used and add them to the `renv.lock` file.

However, in science, we want to be more explicit rather than implicit.
The way `{renv}` explicitly adds to the `renv.lock` file is by only
scanning the `DESCRIPTION` file. While we have already selected to use
explicit declaration when initilizing `{renv}`, this option is specific
to our R instance. In order for `{renv}` to always work explicitly, we
need to set a project option for it. This option needs to be added to
the project's `.Rprofile` file.

We also want to use the `{pak}` package to install R packages instead of
`install.packages()` because it is much more powerful and faster. We can
tell `{renv}` to use `{pak}` instead by adding an option, which we will
be doing anyway.

We can quickly open this with:

```{r open-rprofile-first-time}
#| eval: false
usethis::edit_r_profile("project")
```

Next, at the top of the file, add this code:

```{r snapshot-settings}
#| eval: false
options(
  renv.settings.snapshot.type = "explicit",
  renv.config.auto.snapshot = TRUE,
  renv.config.pak.enabled = TRUE
)
```

```{r purl-only-snapshot-settings}
#| echo: false
#| eval: false
#| purl: true
# Need this for below code.
write_to_file(
  'options(renv.settings.snapshot.type = "explicit", renv.config.auto.snapshot = TRUE, renv.config.pak.enabled = TRUE)',
  ".Rprofile",
  append = TRUE
)
git_ci(".Rprofile", "Add renv snapshot settings to profile.")
git_ci(c("renv.lock", "renv"), "Save lock file.")
```

These three options make it so that whenever you add a package with
`usethis::use_package()` or `pak::pak()`, `{renv}` will always run
`renv::snapshot()` and the snapshot explicitly only look at the
`DESCRIPTION` file. Let's restart the R session so that the `.Rprofile`
changes get activated.

Now our `{renv}` workflow will largely be automated for us, as long as
we do `usethis::use_package()`. If we ever ever return to a project or
collaborate on a project that uses `{renv}`, we can install all the
necessary packages with:

```{r renv-restore}
#| eval: false
renv::restore()
```

And if we need to update packages, we use:

```{r renv-update}
#| eval: false
renv::update()
```

*Sometimes*, working with `{renv}` can get annoying and you just need to
finish working on a task. If that's the case, you can always do
`renv::deactivate()` to stop using `{renv}` and `renv::activate()` to
reactivate it.

**Let's commit the changes made to the Git history** with
{{< var keybind.git >}}.

::: callout-note
## Reading task: \~5 minutes

It's surprising how many issues can come up, from a reproducibility
perspective, when it comes to managing package dependencies. You think
something works well on your computer, but when you create a "virtual
environment" like you do when using `{renv}`, you realize it might not
work as well on other computers.

When this happens, there are several functions you can use to help debug
the situation.

`renv::diagnostics()`

:   List a *lot* of diagnostic information to look over. Sometimes its
    too much, but can help figure out what's going on.

`renv::clean()`

:   Installing packages can sometimes lead to issues in the files of the
    installed packages themselves or even left over temporary files.
    This function tries to clean up these issues for you. It can also
    clean up any unused packages

`renv::repair()`

:   Because of the way `{renv}` works, the connection to where an
    installed package is actually found can get broken. So this function
    tries to fix that and reinstall these broken packages.

`renv::rebuild()`

:   The last resort, use this to reinstall everything from scratch.
:::

## Two types of dependencies

::: {.callout-note appearance="minimal" collapse="true"}
## Instructor note

Verbally explain the content of the next few paragraphs (don't need to
show the website on the projector).
:::

When you work on a research project that involves data analysis, you
likely use packages in two different ways:

1.  Packages that directly contribute to data wrangling, analysis,
    plotting, and making the manuscript. These types of packages are
    generally called "build" or "deploy" dependencies. A package like
    `{dplyr}` or `{tidyr}` would be build dependencies, since you use
    them for processing data.
2.  Packages that assist you in doing your work but aren't directly used
    for data analysis. These types of packages would be called
    "workflow" or "development" dependencies. `{renv}` would be
    considered a workflow dependency.

A good way to determine if a package is a *build dependency* for your
project is by seeing if you write and use functions from the package
within an R script that does something to the data or analysis. If you
*only* ever use functions from the package in the Console, than it is
likely a *workflow dependency*.

The way you add these packages is different depending on the type it is.
For build dependencies, we use the function we've already used before:
`usethis::use_package("packagename")`. For workflow dependencies, it's
the same function, but with a small difference. BUT! Before we cover it,
let's add a setting to our `.Rprofile` to make our life a bit easier. We
will be using `{usethis}` functions many times throughout the course, so
a simple quality-of-life fix is to make it so we don't always have to do
`usethis::`. Thankfully, there is a function that can help. Create and
open the project `.Rprofile` with `usethis::edit_r_profile()`.

```{r add-usethis-rprofile}
#| eval: false
usethis::edit_r_profile("project")
```

Then copy and paste this code into the `.Rprofile`:

``` r
if (interactive()) {
  suppressMessages(require(usethis))
}
```

```{r purl-only-usethis-rprofile}
#| echo: false
#| eval: false
#| purl: true
write_to_file("if (interactive()) {suppressMessages(require(usethis))}",
  ".Rprofile",
  append = TRUE
)
git_ci(".Rprofile", "Add usethis settings to profile.")
```

Let's restart R with {{< var keybind.restart-r >}} before using
`use_package()` to add `{renv}` as a workflow dependency.

```{r suggests-dep}
#| eval: false
#| purl: true
use_package("renv", "suggests")
```

```{r purl-only-update-desc-with-renv}
#| eval: false
#| echo: false
#| purl: true
# Admin: Restart R session first.
git_ci("DESCRIPTION", "Add renv as a suggests dependency.")
```

Open the Git interface and see that under `Suggests:` in the
`DESCRIPTION` file is `renv`. Let's commit these changes, as well as the
changes to `.Rprofile`, with {{< var keybind.git >}}.

::: {.callout-note collapse="true"}
## Instructor note

For the information block below, mention it to the learners but you
don't need to go over it. Especially mention the second part of the tip.
:::

::: {.callout-tip appearance="default"}
When you come back to a project after a few months or if you start
collaborating on a project, usually `renv::restore()` will be enough to
install all types of dependencies. Sometimes though, workflow
dependencies might not get installed. As long as they are tracked in the
`DESCRIPTION` file though, you can force installation of them with:

```{r install-deps}
#| eval: false
renv::install()
```

Depending on your operating system (Windows, MacOS, or Linux), using
`use_package()` might actually install the package again, even though
you already have it. In that case, you can use
`renv::install("packagename")` first, before than using `use_package()`.
:::

## Exercise: Connect your project to GitHub

> Time: \~25 minutes.

Since we will eventually connect our project Git repository to GitHub to
display a website as well as to practice the workflow, we'll connect our
project to GitHub right now.

Let's complete these tasks to connect to GitHub.

1.  First, commit the latest changes to the Git history with
    {{< var keybind.git >}}.

2.  If you haven't yet, please create a
    [GitHub](https://github.com/join) account.

3.  Add the `{gitcreds}` package as a workflow dependency. You'll need
    it for the next item

    ``` r
    use_package("gitcreds", "suggests")
    ```

4.  Read through and complete the tasks in the [Connect to GitHub
    Guide](https://guides.rostools.org/connect-github).

5.  Check your GitHub to make sure the project repository has been
    uploaded to it.

## Automatically adhere to a style guide

::: callout-note
## Reading task: \~5 minutes

We've covered how to adhere to a style guide in both the introductory as
well as the intermediate course, because it is such a useful and
powerful tool to easily write more readable code. We'll expand on it in
this course because it fits precisely with the theme of collaboration.
That's because, when you're working on your own and not needing to worry
about anyone seeing your code, there's a natural temptation to write
your code like you might write notes to yourself... scribbled and
scrawled down quickly. But when working with others, how it looks can
greatly impact how quickly and easily others are able to read and
interpret your code. Multiply all the collaborators on a project with
this natural temptation for quick (and sloppy) coding, you can imagine
how easy it is for code to massively "drift" towards being poorly
formatted, especially when deadlines are close.

That's when "linters" or "stylers" (types of "static code analysis"
tools) become very useful. They will scan your code for common mistakes
or syntax problems and either list them out for you to fix or fix them
for you automatically. Linters are great when you are collaborating on a
project with collaborators who are not as experienced in writing code or
who only occasionally contribute so don't know the workflow culture of
your project. In this way, you might want to have automatic
linting/styling checks that are independent of you having to run them
yourself. This is where the `{styler}` package comes in!

Since we will use it for the project as a workflow dependency, let's add
it to the `DESCRIPTION` file.

```{r add-styler-dep}
#| eval: false
#| purl: true
use_package("styler", "suggests")
```

```{r purl-only-update-desc-with-styler}
#| eval: false
#| echo: false
#| purl: true
git_ci("DESCRIPTION", "Add styler as a suggests dependency.")
```
:::

There are only a few functions in `{styler}` that we need to use. The
first is to style a single file by using `styler::style_file()`.
However, an easier function is the "style active file" RStudio addin.
Usually we would only need to style the file that we are actually
working on, We can do that through the Command Palette
({{< var keybind.palette >}}) and typing "style file", which should show
the "Style active file" option. This is what we will do frequently
throughout the course.

Let's try it out. While inside the `data-raw/nmr-omics.R` file, use
{{< var keybind.styler >}} to style the file. There won't be any changes
since the file is already tidy.

::: {.callout-note appearance="default"}
You will probably be asked to install something, click "Yes".
:::

If you wanted to run `{styler}` on all the files, we can use:

```{r style-dir}
#| eval: false
styler::style_dir()
```

The thing to note, though, is that `{styler}` isn't perfect, so you
might sometimes have to manually fix run the reformatting (with the
Command Palette {{< var keybind.palette >}} then "reformat").

::: {.callout-note collapse="true"}
## Instructor note

Mention the callout block below, but don't go into it at all.
:::

::: {.callout-tip appearance="default"}
You might be used to using 4 spaces for tabs instead of 2. The tidyverse
style uses 2, so the default option in `{styler}`

```{r styler-options-4-spaces}
options(
  styler.addins_style_transformer = "styler::tidyverse_style(indent_by = 4)"
)
```
:::

## Styling Markdown files

For multi-person collaborative projects, having some type of code
styling and checker can really help with standardizing how the code
looks, which ultimately will make it easier to read each other's code
contributions.

But what about for Markdown files? While there isn't a package or
function (yet) that styles the Markdown files, RStudio does have an
option in their Tools to format Markdown into a "canonical form". The
reason for this option is because they added a "visual editor mode" to
writing R Markdown files (which is great if you are more comfortable
with apps like Word). Let's test out this option. First, let's make sure
everything has been committed to the Git history with
{{< var keybind.git >}}.

::: {.callout-warning appearance="default"}
Use this option *only if* you have your project under Git version
control, since it will directly modify and overwrite the contents of the
entire file.
:::

There are two ways of doing this:

1.  Going into
    [`Tools -> Project Options -> R Markdown`](https://rstudio.github.io/visual-markdown-editing/options.html#project-options)
    and changing the options "Automatic text wrapping" to "column" (with
    the default 72 width value) and "Write canonical visual mode
    markdown" to "true".
2.  Or setting [YAML
    options](/https://rstudio.github.io/visual-markdown-editing/markdown.html#writer-options)
    in either the project-level `_quarto.yml` file (we will cover this
    in @sec-build-website) or at the file-level in the YAML header.

For right now, we will do the file-level YAML settings. Open the
`doc/learning.qmd` file and go to the top of the file. Right below the
last `---`, create a new line above it and paste this code in:

``` yaml
editor_options:
  markdown:
    wrap: 72
    canonical: true
```

```{r purl-only-editor-yaml-metadata}
#| eval: false
#| echo: false
#| purl: true
editor_options <-
  "editor_options:
  markdown:
    wrap: 72
    canonical: true"
readLines("doc/learning.qmd") |>
  append(editor_options, after = 1) |>
  writeLines("doc/learning.qmd")
git_ci("doc/learning.qmd", "Add editor options to qmd file.")
```

Now, when you save your file, RStudio should automatically reformat the
Markdown into a standardized format. If you want to switch to using the
Visual Mode, use {{< var keybind.visual_edit >}} or the "Visual" button
at the top of the Source Pane beside the bolding and italicizing
buttons.

The instructors won't be using the Visual Mode during the course,
however you are welcome to. We will be using the "canonical" markdown
mode though.

Let's test it out. While in the `doc/learning.qmd` file, go to the
bottom of the file and type out:

``` markdown
## This is poorly formatted
- Definitely should have an empty space above this list.
- This isn't a list, why not?
```

Save the file. What happens? Lists in Markdown need to have an empty
space above them to work properly (except for when below a header, but
in all other cases it needs a space above). With this canonical mode on,
we can get feedback right away that it isn't right. It gets
automatically fixed by adding that empty space.

``` markdown
## This is poorly formatted

-  Definitely should have an empty space above this list.
-  This isn't a list, why not?
```

Since this mode is on automatically in the `doc/learning.qmd` file, as
we work through the sessions, we'll get lots of experience using it.
We'll also eventually switch so that the whole project uses this mode.

## Exercise: A few small changes to improve your workflow

> Time: \~2 minutes.

There are many options inside the Global Options in RStudio that can
help you work better and faster. There are a few that will help a lot,
especially in this course and with the workflows we are showing:

-   Go into `Tools -> Global Options -> Code -> Saving`.
    -   Under the heading "General", tick on all of those check boxes.
    -   Under the heading "Auto-save", tick on both those check boxes.

## Exercise: Update the README file, while using canonical markdown mode

> Time: \~10 minutes.

Open up the `README.md` file and copy and paste these YAML metadata to
the top:

``` markdown
---
editor_options:
  markdown:
    wrap: 72
    canonical: true
---
```

Then, start completing the `TODO` items. Save often and watch as the
Markdown gets reformatted. After you are done, commit the changes you
made to the Git history with {{< var keybind.git >}}. Then delete both
`TODO.md`, followed by committing these deletions in the Git history.
Click the "Push" button to push the changes to GitHub.

```{r purl-only-editor-options-to-readme}
#| echo: false
#| eval: false
#| purl: true
editor_yaml <- "
---
editor_options:
  markdown:
    wrap: 72
    canonical: true
---
"
readLines("README.md") |>
  append(editor_yaml, after = 1) |>
  writeLines("README.md")
gert::git_add("README.md")
gert::git_commit("Add editor options to README file.")
fs::file_delete("TODO.md")
gert::git_commit("Don't need this file anymore.")
```

## Add r3 in order to complete the survey

We're at the end of the session and now need to fill in the survey.
There is a function in r3, but r3 isn't installed in our project
library. However, because r3 has a few GitHub packages as dependencies,
we need to tell `{renv}` to authenticate us to GitHub so we can install
as much as we want (GitHub only allows a few unauthenticated installed,
as a security feature against hackers and cyber attacks). Using
`gitcreds::gitcreds_set()` with the GitHub token we created should be
enough for `{renv}` to know what to do. If you encounter an error about
a GitHub API limit, run this code again so you can authenticate yourself
with your token.

```{r gitcreds-set-token}
#| eval: false
gitcreds::gitcreds_set()
```

Then we can install the r3 package into our project with
`renv::install()`. Because the r3 package is on GitHub, we need to tell
`{renv}` about its location by adding `github::` followed by the "user"
(which is `rostools`) and than the repository (`r3`).

```{r install-r3}
#| eval: false
renv::install("github::rostools/r3")
```

Now we can open the feedback survey:

```{r}
#| eval: false
r3::open_feedback_survey()
```

## Summary

-   Track your project package dependencies with `{renv}` and combine it
    with `options()` to automatically make snapshots so you can use the
    `use_package()` function.
-   Install the necessary dependencies with `renv::restore()` or
    `renv::install()`.
-   Follow a style guide by using `{styler}`. Combine with the Command
    Palette ({{< var keybind.palette >}}) to quickly run their functions
    on code you are actively working on.
-   Use RStudio's canonical markdown mode to reformat Markdown into a
    standard format.
