# A general approach to doing statistical analyses {#sec-stats-analyses-basic}

```{r setup}
#| include: false
source(here::here("R/functions.R"))
extract_functions_from_qmd()
source(here::here("R/project-functions.R"))
library(tidyverse)
library(tidymodels)
lipidomics <- read_csv(here::here("data/lipidomics.csv"))
```

Running statistical analyses is a relatively methodical and well-defined
process, even if there is often a lot of trial and error involved.
Sometimes it may feel overwhelming and complicated, which it definitely
can be, but it doesn't have to. By taking a structured approach to
running statistical analyses, you can make it easier on yourself and
feel more in control.

In R, statistical methods are often created and developed by researchers
with little to no training in software development and who often use
them differently. This has some advantages, like having the cutting edge
statistical methods available to us, but has a major disadvantage of
often having to learn a completely different way of running a
statistical analysis, even if it is fairly similar to ones you've used
before. So having a framework for running statistical analyses,
regardless of who created them, can provide that needed structure and
vastly simplify the analysis. This session will be covering a general
framework for running statistical analyses, regardless of the exact
statistical method.

## Learning objectives {#sec-stats-analyses-basic-learning-objectives}

The overall objective for this session is to:

1.  Describe the basic framework underlying most statistical analyses
    and use R to generate statistical results using this framework.

More specific objectives are to:

1.  Describe the general "workflow" and steps involved in stating the
    research question, constructing a model to help answer the question,
    preparing the data to match the requirements of the model, fitting
    it to the data, and finally extracting the results from the fitted
    model.
2.  Categorize the model definition step as a distinct, theory-driven
    step, separate from the data, and use `{parsnip}` functions to help
    with defining your model.
3.  Identify various data transformation techniques and evaluate which
    are good options given the data. Use functions in the `{recipes}`
    package to apply these transformations to your data.
4.  Use the `{broom}` package to extract the model results to later
    present them in graphical format (with `{ggplot2}`).
5.  Continue applying the concepts and functions used from the previous
    sessions.

Specific "anti"-objectives:

-   Will **not** know how to choose and apply the appropriate
    statistical model or test, nor understand any statistical theory,
    nor interpret the statistical results correctly, nor determine the
    relevant data transformations for the statistical approach. What we
    show, we show *only as demonstration purposes only*, they could be
    entirely wrong in how to do them correctly if an expert were to
    review them.

::: {.callout-warning appearance="default"}
We will be making *a lot* of function throughout this session and the
next. This is just a fair warning!
:::

## Exercise: What does a "model" mean?

> Time: \~6 minutes.

In science and especially statistics, we talk a lot about "models". But
what does model actually mean? What different types of definitions can
you think of? Is there a different understanding of model in statistics
compared to other areas?

1.  Take 1 minute to think about your understanding of a model.
2.  Then, over the next 3 minutes, discuss with your neighbour about the
    meaning of "model" and see if you can come to a shared
    understanding.
3.  Finally, over the next 2 minutes, we will share all together what a
    model is in the context of data analysis.

## Theory and "workflow" on statistical modeling

::: {.callout-note collapse="true"}
## Instructor note

Let them read it over than go over it briefly, focusing on what a model
is, that we should create research questions while thinking in the
framework of models, and the general workflow for doing statistical
analyses.
:::

<!-- TODO: Create a revealjs presentation on this? -->

:::: callout-note
## Reading task: \~15 minutes

Almost all fields of research, and definitely more heavily-quantitative
and scientific fields like biomedicine and health, have math and
statistics at the core of taking data to draw inferences or general
observations about the world.

Any time we collect data and need to interpret what it means, we need
statistics. And anytime we want to make inferences about the world from
the data, we need to use statistics to determine the likelihood, or
rather the uncertainty, in those inferences. Statistics is meant to
quantify uncertainty.

How do we quantify uncertainty? By first creating a "theoretical model"
that expresses mathematically our research question. For instance, we
have a theoretical model that outdoor plants grow (non-linearly) with
water and sunlight, but that more sunlight likely means less water (less
rain). While *any* research question could be translated into a
theoretical model, not all theoretical models can be *tested* against
the real world. And that's where the second part comes in: Our
theoretical model needs to be structured in a way that allows us to
measure the items ("parameters") in our model, so that we can build a
mathematical model from the data ("test it in the real world").

```{mermaid fig-model-plant-growth}
%%| label: fig-model-plant-growth
%%| fig-cap: Simple example of a theoretical model of plant growth.
%%| echo: false
%%| eval: true
graph LR
    Sunlight --> Growth
    Water --> Growth
    Sunlight --- Water
```

Great research questions are designed in a way to fit a theoretical
model on measurable parameters, so we can ultimately quantify the
uncertainty in our observations (the data) and in the model. And the
basic simplified math of a statistical model looks mostly the same:

$$y = intercept + x + error $$

So if we use the example of plant growth, it would look like:

$$Growth = Sunlight + Water$$

It's a bit more complicated than this, but this is enough to describe it
for this course. Throughout the rest of the session, we use specific
terms to describe each item in this formula. "Outcome" (also called
"dependent variable") refers to the $y$, "predictor" (or "independent
variable") refers to the $x$. The error and intercept are calculated for
us when we fit the model to the data. The intercept is when x is equal
to zero (the "y-intercept" on a plot). The error is the difference
between what the model estimates and the real value. It plays a role in
quantifying the model's uncertainty.

Considering the mathematical nature of statistical models, there is also
a logic and "workflow" to making these models!

1.  Write a research question, designed in a way that might look like
    the diagram above. Usually this step needs to be revisited, revising
    the question after trying to construct the theoretical model, that
    describes the measurable (and unmeasurable) parameters in the model,
    and vice versa.
2.  Based on the model and the type of measured parameters used
    (continuous or binary), select the best mathematical model "type".
    Nearly all models in statistics start from the base of a linear
    regression (e.g. ANOVA is a special form of regression, t-test is a
    simpler version of ANOVA), so the model "type" will probably be a
    form of regression.
3.  Measure your parameters (in the plant growth example, that might be
    the amount of water given in liters per day, amount of plant growth
    in weight, and amount of sunlight in hrs per day). Usually, this
    measured data need to be processed in a special way to fit the
    specifics of the model, research question, and practices of the
    field.
4.  Fit the data to the theoretical model in order to estimate the
    values ("coefficients") of the model parameters as well as the
    uncertainty in those values.
5.  Extract the values and their uncertainty from the model and present
    them in relation to your research questions.

```{mermaid fig-model-building-workflow}
%%| label: fig-model-building-workflow
%%| fig-cap: Simple schematic of the workflow for conducting statistical analysis.
%%| echo: false
%%| eval: true
graph TD
    A[Research question] --> B((Statistical model))
    B --> A
    B --> C[Data collection]
    C --> D[Data transformation]
    D --> E[Scientific output]
    E --> A
```

::: {.callout-caution appearance="default"}
The entire workflow for building statistical models requires highly
specific domain knowledge on not only the statistics themselves, but
also how the data was collected, what the values mean, what type of
research questions to ask and how to ask them, how to interpret the
results from the models, and how to process the data to fit the question
and model.

For instance, in our `lipidomics` dataset, if we were to actually use
this data, we would need someone familiar with -omic technologies, how
the data are measured, what the values actually mean, how to prepare
them for the modeling, the specific modeling methods used for this
field, and how we would actually interpret the findings. We have
**none** of these things, so very likely we are doing things quite wrong
here. We're only doing this modeling to highlight how to use the R
packages.
:::

Going back to our own `lipidomics` dataset, we need to do the first
step: Creating the question. While we don't have much data, there are a
surprising number of questions we could ask. But we will keep it very
simple, very basic, and very exploratory.

1.  What is the estimated relationship of each metabolite with T1D
    compared to the controls, adjusting for the influence of age and
    gender?

Next, because we are working within a "reproducible analysis" framework
specifically with the use of `{targets}`, let's convert this question
into outputs to include as pipeline targets, along with a basic idea for
the final functions that will make up these targets and their inputs and
outputs. These targets will probably be quite different by the end, but
it's a good start to think about what it should look like in the end.

-   All results for estimated relationships (in case we want to use it
    for other output)
-   Plot of statistical estimate for each relationship

Potential function names might be:

-   `calculate_estimates()`
-   `plot_estimates()`

```{mermaid fig-model-possible-targets}
%%| label: fig-model-possible-targets
%%| fig-cap: Potential inputs, outputs, and functions for the targets pipeline.
%%| echo: false
%%| eval: true
graph TB
    lipidomics -- "calculate_estimates()" --> model_est[Model estimates]
    model_est -- "plot_estimates()" --> plot_est[Plot of estimates]
    plot_est -- "tar_read()" --> qmd[Quarto]
```
::::

::: {.callout-note collapse="true"}
## Instructor note

A few things to repeat and reinforce:

1.  The workflow of the image and that it all starts with the research
    question.
2.  The fact that almost all statistical methods are basically special
    forms of linear regression.
3.  That this model creation stage requires a variety of domain
    expertise, not just statistical expertise.

Also repeat the question to ask, the outputs, as well as the functions
and their flow that we can translate into the `{targets}` pipeline.
:::

## Defining the model

::: {.callout-note collapse="true"}
## Instructor note

Let them read it first and then briefly verbally walk through this
section, describing the theoretical model both graphically and
mathematically. Go through why we use `{tidymodels}` rather than other
approaches.
:::

::: callout-note
## Reading task: \~10 minutes

Now that we've talked about the workflow around making models and have
already written out some research questions, let's make a basic,
graphical theoretical model:

```{mermaid fig-model-research-question}
%%| label: fig-model-research-question
%%| fig-cap: A simple theoretical model of the research question about T1D status.
%%| echo: false
%%| eval: true
graph TB
    Metabolite --> T1D
    Age & Gender --> T1D & Metabolite
```

Or mathematically:

$$T1D = metabolite + age + gender$$

So, T1D status (or `class` in the `lipidomics` dataset) is our
**outcome** and the individual metabolite, age, and gender are our
**predictors**. Technically, age and gender would be "confounders" or
"covariates", since we include them only because we think they influence
the relationship between the metabolite and T1D.

If we convert the formula into a form with the variables we have in the
dataset as well as selecting only one metabolite for now (the
cholesterol metabolite, which we add "metabolite" to differentiate it
from other potential variables), it would be:

$$class = metabolite\_cholesterol + age + gender$$

Now that we have a theoretical model, we need to choose our model type.
Since T1D is binary (either you have it or you don't), the most likely
choice is logistic regression, which requires a binary outcome variable.
So we have the theoretical model and the type of model to use - how do
we express this as code in R? There are many ways of doing the same
thing in R, but some are a bit easier than others. One such approach,
that is quite generic and fits with the ideals of the `{tidyverse}`, is
a similar universe of packages called the `{tidymodels}`.

Why do we teach `{tidymodels}`? Because they are built by software
developers, employed by Posit (who also employs the people who build the
`{tidyverse}` and RStudio), and they have a strong reputation for
writing good documentation. Plus, the `{tidymodels}` set of packages
also make creating and using models quite generic, so by teaching you
these sets of tools, you can relatively easily change the model type, or
how you process the data, or other specifications without having to
learn a whole new package or set of tools.

The reason `{tidymodels}` can do that is because they designed it in a
way that makes a clear separation in the components of the model
building workflow that was described above, through the use of specific
packages for each component.

| Package | Description |
|------------------------------------|------------------------------------|
| `{parsnip}` | Model definition, such as type (e.g. `linear_reg()`) and "engine" (e.g. `glm()`). |
| `{recipes}` | Model-specific data transformations, such as removing missing values, or standardizing the data. |
| `{workflows}` | Combining model definition, data, and transformations to calculate the estimates and uncertainty. |

: Core packages within `{tidymodels}`.

We'll start with the `{parsnip}` package. Functions in this package are
used to set the details of the model you want to use. Specifically,
[functions](https://parsnip.tidymodels.org/reference/index.html#models)
to indicate the model *"type"* (e.g. linear regression) and the
`set_engines()` function to determine the "engine" to run the type
(which R-based algorithm to use, like `glm()` compared to `lm()`). Check
out the
[Examples](https://parsnip.tidymodels.org/articles/Examples.html) page
for code you might use depending on the model you want. The most
commonly used model types would be `linear_reg()`, `logistic_reg()`, and
`multinom_reg()`.
:::

Let's switch to working in the `doc/learning.qmd` file to create those
logistic regression models. Since we've already created some items, such
as the lipidomics data, in the `{targets}` pipeline, we'll need to tell
the Quarto file where to find this "store" of items, and import the data
using `tar_read()`. We also need to add `library(tidymodels)` to the
`setup` code chunk. Copy and paste this code chunk below into the Quarto
file.

```` {.markdown filename="doc/learning.qmd"}
```{{r setup}}
targets::tar_config_set(store = here::here("_targets"))
library(tidyverse)
library(targets)
library(tidymodels)
source(here::here("R/functions.R"))
lipidomics <- tar_read(lipidomics)
```
````

Since we will be using `{tidymodels}`, we need to install it, as well as
explicitly add the `{parsnip}`, `{recipes}`, and `{workflows}` packages.
Like `{tidyverse}`, we need to set `{tidymodels}` differently because it
is a "meta-package". We might need to force installing it with
`pak::pak("tidymodels")`.

```{r tidymodels-to-deps}
#| filename: Console
#| purl: true
#| eval: false
use_package("tidymodels", "depends")
# pak::pak("tidymodels")
use_package("parsnip")
use_package("recipes")
use_package("workflows")
```

Before continuing, let's **commit** the changes to the Git history with
{{< var keybind.git >}}. Next, in the `doc/learning.qmd` file, on the
bottom of the document create a new header and code chunk:

```` {.markdown filename="doc/learning.qmd"}
## Building the model

```{{r}}

```
````

In the new code chunk, we will set up the model specs:

```{r logistic-reg-specs}
#| filename: "doc/learning.qmd"
log_reg_specs <- logistic_reg() |>
  set_engine("glm")
log_reg_specs
```

Running this on it's own doesn't show much, as you can see. But we've
now set the model we want to use.

## Data transformations specific to modeling

Setting the model type was pretty easy right? The more difficult part
comes next with the data transformations. `{recipes}` functions are
almost entirely used to apply transformations that a model might
specifically need, like mean-centering, removing missing values, and
other aspects of data processing.

Let's consider our `lipidomics` dataset. In order for us to start our
statistical analysis, we need the data to be structured in a certain way
to be able to smoothly use it as input in our model. We have at least
three easy observations on necessary transformations of the data, two of
which can be fixed with a single `{tidyr}` function, while the third one
can be fixed with `{recipes}`. Can you spot them?

```{r print-lipidomics-data}
#| column: page-inset-right
#| filename: Console
lipidomics
```

::: {.callout-note collapse="true"}
## Instructor note

Ask them if they can spot these differences. Give them a few minutes to
think and respond.
:::

The first observation isn't always an issue and depends heavily on the
model type you use. Since we are using logistic regression, the model
assumes that each row is an individual person. But our data is in the
long format, so each person has multiple rows. The second observation is
that there seems to be a data input error, since there are multiple
`Cholesterol` values, while all other metabolites only have one:

```{r too-many-cholesterols}
#| filename: "doc/learning.qmd"
lipidomics |>
  count(code, metabolite) |>
  filter(n > 1)
```

We can fix both the long format and multiple cholesterol issues by using
`tidyr::pivot_wider()`. Before we do, the last issue is that each
metabolite has quite large differences in the values and ranges of data.
Again, whether this is an issue depends on what we want to do, but in
our research question we want to know how each metabolite influences
T1D. In order to best interpret the results and compare across
metabolites, we should ideally have all the metabolites with a similar
range and distribution of values.

Let's fix the first two issues first. While we probably only need to use
`pivot_wider()`, we should tidy up the metabolite names first so they
make better column names. We do that by combining `mutate()` with
`snakecase::to_snake_case()`. In the `doc/learning.qmd` file, we rename
the metabolite names before using `pivot_wider()`. Since we want an easy
way of identifying columns that are metabolites, we will add a
`"metabolite_"` prefix using the argument `names_prefix`. To actually
fix the multiple cholesterol issue, we should look more into the data
documentation or contact the authors. But for this course, we will merge
the values by calculating a mean before pivoting. We do this by setting
the `values_fn` with `mean`.

```{r lipidomic-to-wider}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
lipidomics_wide <- lipidomics |>
  mutate(metabolite = snakecase::to_snake_case(metabolite)) |>
  pivot_wider(
    names_from = metabolite,
    values_from = value,
    values_fn = mean,
    names_prefix = "metabolite_"
  )
lipidomics_wide
```

Since we're using a function-oriented workflow and since we will be
using this code again later on, let's convert both the "metabolite to
snakecase" and "pivot to wider" code into their own functions, before
moving them over into the `R/functions.R` file.

```{r first-snakecase-fn}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
column_values_to_snake_case <- function(data) {
  data |>
    dplyr::mutate(metabolite = snakecase::to_snake_case(metabolite))
}
lipidomics |>
  column_values_to_snake_case()
```

This on its own should work. *However*, the column we want to change
might not always be called `metabolite`, or we might want to change it
later. So, to make this function a bit more generic, we can use
something called "curly-curly" (it looks like `{{}}` when used) and
"non-standard evaluation" (NSE).

::: callout-note
## Reading task: \~10 minutes

When you write your own functions that make use of functions in the
`{tidyverse}`, you may eventually encounter an error that might not be
very easy to figure out. Here's a very simple example using `select()`,
where one of your function's arguments is to select columns:

```{r test-nse, error=TRUE}
test_nse <- function(data, columns) {
  data |>
    dplyr::select(columns)
}

lipidomics |>
  test_nse(class)
```

The error occurs because of something called "[non-standard
evaluation](http://adv-r.had.co.nz/Computing-on-the-language.html)" (or
NSE). NSE is a major feature of R and is used quite a lot throughout R.
NSE is used a lot in the `{tidyverse}` packages. It's one of the first
things computer scientists complain about when they use R, because it is
not a common thing in other programming languages. But NSE is what
allows you to use formulas (e.g. `y ~ x + x2` in modeling, which we will
show shortly) or allows you to type out `select(class, age)` or
`library(purrr)`. In "standard evaluation", these would instead be
`select("Gender", "BMI")` or `library("purrr")`. So NSE gives
flexibility and ease of use for the user (we don't have to type quotes
every time) when doing data analysis, but can give some headaches when
programming in R, like when making functions. There's more detail about
this on the [dplyr
website](https://dplyr.tidyverse.org/articles/programming.html#warm-up),
which lists some options to handle NSE while programming. The easiest
approach is to wrap the argument with "curly-curly" (`{{}}`).

```{r test-nse-fixed}
test_nse <- function(data, columns) {
  data |>
    dplyr::select({{ columns }})
}

lipidomics |>
  test_nse(class)

lipidomics |>
  test_nse(c(class, age))
```
:::

::: {.callout-note collapse="true"}
## Instructor note

You don't need to go over what they read, you can continue with making
the function below. Unless learners have some questions.
:::

We can use curly-curly (combined with `across()`) to apply
`snakecase::to_snake_case()` to columns of our choice.

```{r second-snakecase-fn}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
column_values_to_snake_case <- function(data, columns) {
  data |>
    dplyr::mutate(dplyr::across({{ columns }}, snakecase::to_snake_case))
}

lipidomics |>
  column_values_to_snake_case(metabolite)
```

Move this new function over into the `R/functions.R` file, add Roxygen
documentation with {{< var keybind.roxygen >}}, style using
{{< var keybind.styler >}}, `source()` the modified `R/functions.R` file
with {{< var keybind.source >}}

```{r new-function-column-values-to-snakecase}
#| filename: "R/functions.R"
#' Convert a column's character values to snakecase format.
#'
#' @param data The lipidomics dataset.
#' @param columns The column you want to convert into the snakecase format.
#'
#' @return A data frame.
#'
column_values_to_snake_case <- function(data, columns) {
  data |>
    dplyr::mutate(dplyr::across({{ columns }}, snakecase::to_snake_case))
}
```

Now add the new function above the `pivot_wider()` code in the
`doc/learning.qmd` file.

```{r snakecase-function-before-pivot-wider}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
lipidomics_wide <- lipidomics |>
  column_values_to_snake_case(metabolite) |>
  mutate(metabolite = snakecase::to_snake_case(metabolite)) |>
  pivot_wider(
    names_from = metabolite,
    values_from = value,
    values_fn = mean,
    names_prefix = "metabolite_"
  )
lipidomics_wide
```

## Exercise: Convert the pivot code into a function

> Time: \~10 minutes.

Just like with the `mutate()`, take the `pivot_wider()` code and convert
it into a new function.

1.  Name the new function `metabolites_to_wider`.
2.  Include one argument in the new `function()`: `data`.
3.  Use `data |>` at the beginning, like we did with the
    `column_values_to_snake_case()`.
4.  Use `tidyr::` before the `pivot_wider()` function.
5.  Add the Roxygen documentation with {{< var keybind.roxygen >}}.
6.  Move the function into the `R/functions.R` file.
7.  Replace the code in the `doc/learning.qmd` file to make use of the
    new functions.

```{r solution-new-function-metabolite-to-wider}
#| eval: true
#| code-fold: true
#| code-summary: "**Click for the solution**. Only click if you are struggling or are out of time."
#' Convert the metabolite long format into a wider one.
#'
#' @param data The lipidomics dataset.
#'
#' @return A wide data frame.
#'
metabolites_to_wider <- function(data) {
  data |>
    tidyr::pivot_wider(
      names_from = metabolite,
      values_from = value,
      values_fn = mean,
      names_prefix = "metabolite_"
    )
}
```

## Using recipes to manage transformations

<!-- TODO: convert this next few paragraphs to reading exercise? -->

::: {.callout-note collapse="true"}
## Instructor note

Verbally talk through the next few paragraphs, no code-along yet.
:::

We've used `{dplyr}` and `{tidyr}` to start fixing some of the issues
with the data. But we still have the third issue: How to make the
results between metabolites comparable. That's where we use `{recipes}`.

The first function is `recipe()` and it takes two forms: with or without
a formula. Remember the model formula we mentioned previously? Well,
here is where we can use it to tell `{recipes}` about the model formula
we intend to use so it knows on what variables to apply the chosen
transformations.

For a few reasons that will be clearer later, we won't ultimately use
the formula form of `recipe()`, but will show how it works:

```{r recipes-with-formula}
recipe(class ~ metabolite_cholesterol + age + gender, data = lipidomics_wide)
```

The alternative approach is to set "roles" using `update_roles()`.
Instead of using a formula and letting `recipe()` infer the outcome and
predictors, we can explicitly select which variables are which. This has
some nice features that we will use later on.

```{r recipes-without-formula}
#| filename: "doc/learning.qmd"
recipe(lipidomics_wide) |>
  update_role(metabolite_cholesterol, age, gender, new_role = "predictor") |>
  update_role(class, new_role = "outcome")
```

::: {.callout-note collapse="true"}
## Instructor note

Describe this next paragraph by switching to this website and showing
the list of some of the steps available in the code chunk below.
:::

The next "step" is to select a transformation function. There are many
`step_*` functions (some shown below) that are available in `{recipes}`
and which one you decide to use depends heavily on your data and your
research question. So take your time thinking through which
transformations you actually need for your analysis, rather than quickly
using one or more of them.

```{r list-step-transforms}
#| eval: false
recipes::step_log()
recipes::step_scale()
recipes::step_normalize()
recipes::step_center()
recipes::step_sqrt()
```

::: {.callout-tip appearance="default"}
There are so many useful transformation functions available. For
instance, if you often have to impute data, there are functions for
that. You can check them out in the Console by typing
`recipes::step_impute_` then hit the Tab key to see a list of them. Or,
if you have some missing values, there's also the
`recipes::step_naomit()`.
:::

There are many transformations we could use for the `lipidomics`
dataset, but we will use `step_normalize()` for this course. This
function is useful because it makes each variable centered to zero and a
value of 1 unit is translated to 1 standard deviation of the original
distribution. This means we can more easily compare values between
variables. We can add this to the end of the recipe:

```{r recipes-with-step-normalize}
#| filename: "doc/learning.qmd"
recipe(lipidomics_wide) |>
  update_role(metabolite_cholesterol, age, gender, new_role = "predictor") |>
  update_role(class, new_role = "outcome") |>
  step_normalize(starts_with("metabolite_"))
```

Next thing to do is convert this into a function, using the same
workflow we've been using (which means this needs to be in the
`R/functions.R` script). We'll also use the curly-curly again, since we
might use a different metabolite later. Note, when adding all the
`packagename::` to each function, the `starts_with()` function comes
from the `{tidyselect}` package.

```{r new-function-create-recipe-spec}
#| filename: "R/functions.R"
#' A transformation recipe to pre-process the data.
#'
#' @param data The lipidomics dataset.
#' @param metabolite_variable The column of the metabolite variable.
#'
#' @return
#'
create_recipe_spec <- function(data, metabolite_variable) {
  recipes::recipe(data) |>
    recipes::update_role({{ metabolite_variable }}, age, gender, new_role = "predictor") |>
    recipes::update_role(class, new_role = "outcome") |>
    recipes::step_normalize(tidyselect::starts_with("metabolite_"))
}
```

And test it out:

```{r use-create-recipe-specs-fn}
#| filename: "doc/learning.qmd"
#| column: page-inset-right
recipe_specs <- lipidomics_wide |>
  create_recipe_spec(metabolite_cholesterol)
recipe_specs
```

Style using {{< var keybind.styler >}}, then commit the changes made to
the Git history with {{< var keybind.git >}}.

## Fitting the model by combining the recipe, model definition, and data {#sec-fitting-model}

We've now defined the model we want to use and created a transformation
`{recipes}` specification. Now we can start putting them together and
finally fit them to the data. This is done with the `{workflows}`
package.

Why use this package, rather than simply run the statistical analysis
and process the data as normal? When running multiple models (like we
will do in the next section) that may require different data structures,
the data transformation steps have to happen *right before* the data is
fit to the model and need to be done on exactly the data used by the
model. So if we have one data frame that we run multiple models on, but
the transformation happens to the whole data frame, we could end up with
issues due to how the transformations were applied. The `{workflows}`
package keeps track of those things for us, so we can focus on the
higher level thinking rather than on the small details of running the
models.

The `{workflows}` package has a few main functions for combining the
recipe with the model specs, as well as several for updating an existing
workflow (which might be useful if you need to run many models of
slightly different types). All model workflows need to start with
`workflow()`, followed by two main functions: `add_model()` and
`add_recipe()`. Can you guess what they do?

```{r use-workflow-for-model}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
workflow() |>
  add_model(log_reg_specs) |>
  add_recipe(recipe_specs)
```

While this code is already pretty concise, let's convert it into a
function to make it simplified. We'll use the same function-oriented
workflow that we've used before, where the function should ultimately be
inside the `R/functions.R` file.

```{r new-function-create-model-workflow}
#| filename: "R/functions.R"
#' Create a workflow object of the model and transformations.
#'
#' @param model_specs The model specs
#' @param recipe_specs The recipe specs
#'
#' @return A workflow object
#'
create_model_workflow <- function(model_specs, recipe_specs) {
  workflows::workflow() |>
    workflows::add_model(model_specs) |>
    workflows::add_recipe(recipe_specs)
}
```

<!-- TODO: image showing pieces at play that we need to eventually fit into targets/questions -->

Instead of using the previously created objects, let's start the model
creation from scratch:

```{r full-model-workflow-from-almost-scratch}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
model_workflow <- create_model_workflow(
  logistic_reg() |>
    set_engine("glm"),
  lipidomics_wide |>
    create_recipe_spec(metabolite_cholesterol)
)
model_workflow
```

Now, we can do the final thing: Fitting the data to the model with
`fit()`! :raised_hands:

```{r fit-model-workflow-to-data}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
fitted_model <- model_workflow |>
  fit(lipidomics_wide)
fitted_model
```

This gives us a lot of information, but what we are mostly interested in
is the model estimates themselves. While this `fitted_model` object
contains a lot additional information inside, `{workflows}` thankfully
has a function to extract the information we want. In this case, it is
the `extract_fit_parsnip()` function.

```{r extract-model-fit}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
fitted_model |>
  extract_fit_parsnip()
```

To get this information in a tidier format, we use another function:
`tidy()`. This function comes from the `{broom}` package, which is part
of the `{tidymodels}`. But we should explicitly add it to the
dependencies:

```{r broom-to-deps}
#| purl: true
#| eval: false
#| filename: Console
use_package("broom")
```

Then, we add the `tidy()` function to our model using the `|>` pipe.
Since we are using a logistic regression model, we need to consider how
we want the estimates to be presented, probably depending on how we want
to visualize our results. If we set `exponentiate = TRUE` in `tidy()`,
the output estimates will be odds ratios, if we set
`exponentiate = FALSE`, we will get the log odds ratios or the beta
coefficient. Here we choose `exponentiate`

```{r tidy-up-model-results}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
fitted_model |>
  extract_fit_parsnip() |>
  tidy(exponentiate = TRUE)
```

We now have a data frame of our model results!

::: {.callout-note collapse="true"}
## Instructor note

Reinforce **not** to worry about interpreting the results, that is not
the aim of this course.
:::

::: callout-important
Another reminder, we **are not** interpreting this results. For this
course, they are **not** important and can distract from the main
purpose.
:::

Like we did with the `workflows()` code that we converted into a
function, we do the same thing here: Make another function (and move it
to `R/functions.R`)! :stuck_out_tongue:

```{r new-function-tidy-model-output}
#| filename: "R/functions.R"
#' Create a tidy output of the model results.
#'
#' @param workflow_fitted_model The model workflow object that has been fitted.
#'
#' @return A data frame.
#'
tidy_model_output <- function(workflow_fitted_model) {
  workflow_fitted_model |>
    workflows::extract_fit_parsnip() |>
    broom::tidy(exponentiate = TRUE)
}
```

Replacing the code in the `doc/learning.qmd` file to use the function.

```{r use-tidy-model-output-fn}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
fitted_model |>
  tidy_model_output()
```

If we revise the code so it is one pipe, it would look like:

```{r single-pipe-model-results}
#| column: page-inset-right
#| filename: "doc/learning.qmd"
create_model_workflow(
  logistic_reg() |>
    set_engine("glm"),
  lipidomics_wide |>
    create_recipe_spec(metabolite_cholesterol)
) |>
  fit(lipidomics_wide) |>
  tidy_model_output()
```

Let's briefly cover what these columns and values mean.

::: {.callout-note collapse="true"}
## Instructor note

If you want, you can go over these details briefly or in more detail,
depending on how comfortable you are. Or you can get them to read it
only.
:::

::: callout-note
## Reading task: \~10 minutes

Let's explain this output a bit, each column at a time:

-   `term`: If you recall the formula
    $class = metabolite + sex + gender$, you'll see all but the `class`
    object there in the column `term`. This column contains all the
    predictor variables, including the intercept (from the original
    model).

-   `estimate`: This column is the "coefficient" linked to the term in
    the model. The final mathematical model here looks like:

    $$ \displaylines{class = Intercept + (metabolite\_estimate \times metabolite\_value) + \\ (gender\_estimate \times gender\_value) + ...}$$

    In our example, we chose to get the odds ratios. In the mathematical
    model above, the estimate is represented as the log odds ratio or
    beta coefficient - the constant value you multiply the value of the
    term with. Interpreting each of these values can be quite tricky and
    can take a surprising amount of time to conceptually break down, so
    we won't do that here, since this isn't a statistics course. The
    only thing you need to understand here is that the `estimate` is the
    value that tells us the *magnitude* of association between the term
    and `class`. This value, along with the `std.error` are the most
    important values we can get from the model and we will be using them
    when presenting the results.

-   `std.error`: This is the uncertainty in the `estimate` value. A
    higher value means there is less certainty in the value of the
    `estimate`.

-   `statistic`: This value is used to, essentially, calculate the
    `p.value`.

-   `p.value`: This is the infamous value we researchers go crazy for
    and think nothing else of. While there is a lot of attention to this
    single value, we tend to give it more attention than warranted. The
    interpretation of the p-value is even more difficult than the
    `estimate` and again, we won't cover this in this course. We won't
    be using this value at all in presenting the results.
:::

Before ending, open the Git interface and commit the changes you made
with {{< var keybind.git >}}. Then push your changes up to GitHub.

## Summary

-   Create research questions that (ideally) are structured in a way to
    mimic how the statistical analysis will be done, preferably in a
    "formula" style like $y = x1 + x2 + ... + error$ and in a diagram
    style with links connecting variables.
-   Statistical analyses, while requiring some trial and error, are
    surprisingly structured in the workflow and steps taken. Use this
    structure to help guide you in completing tasks related to running
    analyses.
-   Use `{parsnip}` functions to define the model you want to use, like
    `logistic_reg()` for logistic regression, and set the computational
    "engine" with `set_engine()`.
-   Use `{recipes}` functions to set up the data transformation steps
    necessary to effectively run the statistical analysis, like adding
    variable "roles" (outcome vs predictor) using `update_roles()` and
    adding transformation steps using any of the dozen different `step_`
    functions.
-   Use `{workflows}` functions to develop an analysis `workflow()` that
    combines the defined model with `add_model()`, the transformation
    steps with `add_recipe()`, and the data with `fit()`.
-   Use `{broom}` to `tidy()` the model output, extracted using
    `extract_fit_parsnip()` to get a data frame of the estimates and
    standard error for the variables in the mel.
