---
execute:
  eval: false
---

```{r setup}
#| eval: true
#| include: false
library(tidyverse)
library(downlit)
```

# Download the workshop data {#sec-download-data}

We're going to use a real world dataset to demonstrate the concepts in
this workshop. We're going to use an openly licensed dataset with
metabolomics data [@Herance2022].

Similar to the intermediate workshop, we will follow the principle of
building a reproducible pipeline, from raw data to the report on the
analysis. We'll also follow the principle of keeping the raw data *raw*
and use code to tidy it up. The ultimate goal is to have a record of how
we *exactly* went from raw data to results in the paper. Unlike the
intermediate workshop where you had to write through each step of the
script and what you needed to do, in this workshop you only need to copy
and paste the code that will download and minimally process the dataset
into an initially workable state. If you want to learn more about how to
take a raw data set and process it into a format that is more suitable
for analysis, check out the [intermediate
workshop](https://r-cubed-intermediate.rostools.org/).

Inside the `data-raw/` folder, we are going to write R code that
downloads a dataset, processes it into a more tidy format, and save it
into the `data/` folder. This is the start of our analysis pipeline.
First step, we need to create the script to do these steps. While in
your `AdvancedR3` R Project, go to the Console pane in RStudio and type
out:

```{r create-data-raw}
#| filename: Console
#| purl: true
usethis::use_data_raw("nmr-omics")
```

What this function does is create a new folder called `data-raw/` and
creates an R script called `nmr-omics.R` in that folder. This is where
we will store the raw, original metabolomics data that we'll get from
the [website](https://zenodo.org/record/6597902). If you go to the
website with the dataset, you'll notice (when you scroll down) that
there are three files: A README file, a metabolomics `.xlsx` file, and a
lipidomics `.xlsx` file. For now, we only want the README file and
lipidomics dataset.

The R script should have opened up for you, otherwise, go into the
`data-raw/` folder and open up the new `nmr-omics.R` script. The first
thing to do is delete all the code in the script. Than, copy and paste
the code below into the script.

```{r, file=here::here('data-raw/nmr-omics.R'), filename="data-raw/nmr-omics.R"}
```

```{r purl-only-paste-processing-code}
#| echo: false
#| purl: true
readLines(
  "https://raw.githubusercontent.com/rostools/r-cubed-advanced/main/data-raw/nmr-omics.R"
) |>
  writeLines(usethis::proj_path("data-raw/nmr-omics.R"))
source(usethis::proj_path("data-raw/nmr-omics.R"))
```

Since this is an advanced workshop, you can run the lines one at a time
and see what they do on your own (or `source()` them all at once). The
comments provided give guidance on what the code is doing and why. In
the end, though, the only important thing is to run all the code and get
the `lipidomics` dataset to be saved as `data/lipidomics.csv`. The
created files should look like this:

```{r}
#| echo: false
fs::dir_tree(regexp = "data(-raw)?")
```

``` text
AdvancedR3
├── data
│   └── lipidomics.csv
└── data-raw
    ├── nmr-omics
    │   ├── README.txt
    │   └── lipidomics.xlsx
    └── nmr-omics.R
```

And when using `readr::read_csv()`, the created dataset in
`data/lipidomics.csv` should look like:

```{r}
#| eval: true
readr::read_csv(here::here("data/lipidomics.csv"))
```

Take a look through the downloaded `data-raw/nmr-omics/README.txt` and
`data-raw/nmr-omics/lipidomics.xlsx` files, as well as the created
`data/lipidomics.csv` to begin better understanding the dataset.

We are using Git to track changes made to files in our project. The
original metabolomics dataset is stored on Zenodo, so we don't actually
need to keep the raw data files in our Git history. So let's tell Git to
ignore the files created in the `data-raw/nmr-omics/` folder. In the
Console, type out the code below. You only need to do this once.

```{r ignore-xlsx-files}
#| filename: Console
#| purl: true
usethis::use_git_ignore("data-raw/nmr-omics/")
```

```{r purl-only-processing-data}
#| echo: false
#| purl: true
git_ci(
  c("data-raw/nmr-omics.R", "data/lipidomics.csv", ".gitignore", "DESCRIPTION"),
  "Script to process omics data into lipidomics csv."
)
```

Next, run this command to check that everything is setup correctly. You
will need to paste this output into the survey at the end.

``` {.r filename="Console"}
r3::check_project_setup_advanced()
```

```
Show folders and files of project:
• Please copy and paste this output into the survey question:
/home/luke/Desktop/AdvancedR3
├── AdvancedR3.Rproj
├── DESCRIPTION
├── R
│   ├── README.md
│   └── functions.R
├── README.md
├── TODO.md
├── data
│   ├── README.md
│   └── lipidomics.csv
├── data-raw
│   ├── README.md
│   ├── nmr-omics
│   │   ├── README.txt
│   │   └── lipidomics.xlsx
│   └── nmr-omics.R
└── doc
    ├── README.md
    ├── learning.qmd
    └── report.Rmd
```

The output should look something a bit like the above text. If it
doesn't, start over by deleting everything in the `data-raw/` folder
*except* for the `data-raw/nmr-omics.R` script and rerunning the script
again. If your output looks a bit like the above, than copy and paste
the output into the survey question at the end.

{{< include /includes/_code-appendix.qmd >}}
