# Schedule

The workshop is structured as a series of participatory live-coding
sessions (instructor and learner coding together) interspersed with
hands-on exercises, all with using a real-world, open dataset. There are
few lectures, only given at the start and end of the workshop. The
general schedule outline is shown in the below table.

```{r schedule-table, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
add_link <- function(text, icon, link) {
  colour <- "#138d75"
  added_link <- paste0("<i class='fa-solid fa-", icon, "' style='color:", colour, "'></i>&nbsp;&nbsp;<a href=\"", link, "\">", text, "</a>")
  if (link == "") {
    added_link <- paste0("<i class='fa-solid fa-", icon, "' style='color:", colour, "'></i>&nbsp;&nbsp;", text)
  }
  return(htmltools::HTML(added_link))
}

schedule <- here::here("includes/schedule.csv") %>%
  readr::read_csv(col_types = "ccccc") %>%
  mutate(across(everything(), ~ if_else(is.na(.x), "", .x))) %>%
  mutate(
    `Session topic` = list(`Session topic`, icon, link) %>%
      purrr::pmap(~ add_link(..1, ..2, ..3)) %>%
      unlist()
  ) %>%
  select(-icon, -link)
```

::: panel-tabset
## Day 1 {.unnumbered}

```{r schedule-table-day-1}
schedule %>%
  filter(Day == "Day 1") %>%
  select(-Day) %>%
  knitr::kable()
```

## Day 2 {.unnumbered}

```{r schedule-table-day-2}
schedule %>%
  filter(Day == "Day 2") %>%
  select(-Day) %>%
  knitr::kable()
```

## Day 3 {.unnumbered}

```{r schedule-table-day-3}
schedule %>%
  filter(Day == "Day 3") %>%
  select(-Day) %>%
  knitr::kable()
```
:::
