---
execute:
  eval: false
---

# Connecting to GitHub {#sec-connect-github}

One of the best ways of sharing your code (or really, any scientific
output, like this website!) is through GitHub. Connecting your Git
repository to GitHub is also a great way to collaborate with others and
to keep a backup of your work.

So, let's connect your project to GitHub! 🥳 Thankfully, `{usethis}` has
a few functions to help us out. Before you can connect your project's
Git repository to GitHub, you need to inform GitHub (authenticate) that
you are the owner of your account.

## Authenticating with GitHub

Any time we do anything on the Internet, there is some risk to having
our information maliciously hacked. This is no different when using
GitHub, so if we can, we should try to be more secure with what we send
across the internet. In fact, most functions that relate to Git or using
GitHub *require* using more secure features in order to work.
`{usethis}` makes this much easier, thankfully, with several functions.
The `{usethis}` website has a really [well written
guide](https://usethis.r-lib.org/articles/articles/git-credentials.html)
on setting it up. Here is a very simplified version of what they
recommend that is relevant for what we are doing in this course.

-   Use personal access tokens (PAT, or simply called a "token") when
    interacting with your GitHub remote repositories while outside of
    the GitHub website (e.g. when using R or `{usethis}`). PAT's are
    like temporary passwords that provide limited access to your GitHub
    account, like being able to read or write to your GitHub
    repositories, but not being able to delete them. They are useful
    because you can easily delete the PAT if you feel someone got access
    to it and prevent it from being used, unlike your own password which
    you would have to manually change if it was stolen.
-   Use a password manager to save the PAT for later use. Using password
    managers is basically a requirement for having secure online
    accounts, because they can generate random and long passwords that
    you don't have to remember.
-   Use packages like `{gitcreds}` to give `{usethis}` access to the PAT
    and to interact with your GitHub repositories. You normally would
    use `{gitcreds}` every time you restart your computer or after a
    certain period of time.

You very likely haven't set up a PAT, but if you are uncertain, you can
always check with:

```{r cs-gh-token-help}
gh_token_help()
```

    • GitHub host: 'https://github.com'
    • Personal access token for 'https://github.com': <unset>
    • To create a personal access token, call `create_github_token()`
    • To store a token for current and future use, call `gitcreds::gitcreds_set()`
    ℹ Read more in the 'Managing Git(Hub) Credentials' article:
      https://usethis.r-lib.org/articles/articles/git-credentials.html
      

The output is saying the token is `<unset>`, which means we need to make
Git and `{usethis}` aware of it. We do that by typing the next function
in the Console to create the token on GitHub (if you haven't created one
already).

```{r cs-create-token}
create_github_token()
```

This function sends us to the GitHub "Generate new token" webpage with
all the necessary settings checked. Set the "Expiry date" to 90 days
(this is a good security feature). Then, click the green button at the
bottom called "Generate token" and you'll have a very long string
generated for you that starts with `ghp_`. **Save this token in your
password manager**. This is the token you will use every time you open
up RStudio and interact with GitHub through R. You do **not** need to
create a new token for each R project or package you make, you only need
to create one after your current token expires (typically every couple
of months), if you've forgotten the token or lost it, or if you've
changed to a new computer. In order for R recognize the token, we need
to use the `{gitcreds}` package as a workflow dependency.

```{r git-creds-deps}
use_package("gitcreds", "suggests")
```

Then, in the Console, run:

```{r cs-gitcreds-set}
gitcreds::gitcreds_set()
```

And then copy and paste your token into the prompt in the Console. This
token usually gets saved for the day (it gets cached), but after
restarting you will need to run the action again. If it asks to replace
an existing one, select the "yes" option. Doing this is a bit like using
the two-factor authentication (2FA) you often have to do when, for
instance, accessing your online bank account or other government
website. In this case, you are telling GitHub (when interacting to it
through RStudio, like uploading and downloading your changes) that you
are who you claim to digitally be.

::: callout-tip
There is another great helper function that runs a lot of checks and
gives some advice when it finds potential problems.

```{r cs-git-sitrep}
git_sitrep()
```

Just to be aware, using this function outputs a *lot* of stuff, most of
which you probably don't even need to know or don't even know what it
means. That's ok, since it is meant as a diagnostic tool.
:::

## Linking your project to GitHub

Now that we can authenticate to GitHub that you are digitally you, we
can connect our project's Git repository to GitHub. If you are new to
Git and GitHub, we strongly recommend starting your first project(s) as
private, in case you accidentally add files you aren't supposed to. It
will also help get you feeling comfortable with using Git and GitHub. In
the Console, run:

```{r add-github}
use_github(private = TRUE)
```

This will take your project and upload it to GitHub. Now, whenever you
use Git and save your changes to the Git history, whenever you "Push"
your changes it will be sent to your project on GitHub. The diagram
below shows how it conceptually looks like:

```{mermaid}
%%| eval: true
graph LR;
    yours(Your local) -->|Push| github(GitHub)
    github -->|Pull| yours
```

The "Your local" is your own computer. Whenever you "push" to GitHub, it
means it will upload your file changes (like synchronizing in Dropbox).
Whenever you "pull" from GitHub, it takes any changes made on GitHub and
downloads them to your "Local" computer.

Why use GitHub? For one, it is one of the most effective ways of
collaborating on a shared project. Hundreds of companies and hundreds of
thousands of workers use Git and services like GitHub to work together
on massive projects. The way collaboration works would conceptually look
like:

```{mermaid}
%%| eval: true
graph LR;
    yours(Your local) -->|Push| github(GitHub)
    github -->|Pull| yours
    others(Collaborator's<br>local) -->|Push| github
    github -->|Pull| others
```

This approach to collaborating makes it much easier to contribute
*directly* (not through emails) to projects and to more easily help
others out with issues.
