# Linting your code for style

<!-- NOTE: This isn't included in the website right now -->

```{r setup}
#| include: false
source(here::here("R/functions.R"))
```

## Learning objectives

1.  Explain the role that following a style guide has on building a
    common approach to reading (and writing) R code, and thus improve
    project-level collaboration.
2.  Use `{lintr}` to programmatically check and apply style guides to
    your project files.

## "Linting" your code

When you're working on your own and not needing to worry about anyone
seeing your code, there's a natural temptation to write your code like
you might write notes to yourself... scribbled and scrawled down
quickly. Even when you are working with others or try to write more
readable code, it still is very easy for "lint"[^linting-1] to
accumulate.

[^linting-1]: You know, the fluff you find in your pockets 😝

That's when "linters" (a type of "static code analysis" tool) become
very useful. A linter will scan your code for common mistakes or syntax
problems and list them out for you to fix. They don't directly fix them
for you, though we will show how to do that later in this session.
Linters are great when you are collaborating on a project with
collaborators who are not as experienced in writing code or who only
occasionally contribute so don't know the workflow culture of your
project. In this way, you might want to have automatic linting checks
(or styling) that are independent of you having to run them yourself.

The main package for doing that is the `{lintr}` package. We'll get a
chance to try out using `{renv}` since we need to install it to use it
in our project.

::: {.callout-note collapse="true"}
## Instructor note

Before continuing, ask the learners which type of dependency they think
`{lintr}` is or should be.
:::

```{r purl-only-renv-install-lintr}
#| eval: false
#| echo: false
#| purl: true
renv::install("lintr")
```

```{r add-lintr-as-dep}
#| eval: false
#| purl: true
use_package("lintr", "suggests")
```

```{r purl-only-update-desc-with-lintr}
#| eval: false
#| echo: false
#| purl: true
git_ci("DESCRIPTION", "Add lintr as a suggests dependency.")
```

Right now we don't have much code to "lint", but as we work through the
course, we will make use of `{lintr}` before committing to Git with
{{< var keybind.git >}}. So, how do we use `{lintr}`? It's actually
pretty easy, since there are only a few functions that do the linting,
even though there are dozens of customizations you can make to
`{lintr}`. The basic is to lint a single file, don't run this though:

```{r lint-one-file}
#| eval: false
lintr::lint("R/functions.R")
```

A more encompassing function checks all files in the project (also,
don't run it yet, we'll do that when we have more code in the project):

```{r lint-project-dir}
#| eval: false
lintr::lint_dir()
```

This might be useful if you want to go over everything to check things.
But, if you are working on a single file and want to do some linting of
that, the better solution is to use the Command Palette
{{< var keybind.palette >}} and type "lint". The first option is to lint
the file you are working on. Let's open up the `data-raw/nmr-omics.R`
script and try it out. There *should* be at least one note that says:

``` markdown
Lines should not be more than 80 characters.
```

::: {.callout-note collapse="true"}
## Instructor note

If there is no note, check if the participants have the `# nolint` at
the end of the line with the second `download.file()`. To show what
happens, get them to remove it, run `lintr::lint_dir()` again, and then
bring it back.
:::

Keep in mind, that not all notes need to be changed. A linter is meant
to be used as a guideline, not as a rule. It helps you to follow best
practices and have some consistency between collaborators, but not to
stress about making sure all notes are addressed. We can suppress lint
warnings by adding a `# nolint` comment.

## Summary

-   Follow a style guide by using `{lintr}`. Combine with the Command
    Palette with {{< var keybind.palette >}} to quickly run their
    functions on code you are actively working on.
