# Linting your code for style

```{r setup}
#| include: false
source(here::here("R/functions.R"))
```

## Learning objectives

1.  Explain the role that following a style guide has on building a
    common approach to reading (and writing) R code, and thus improve
    project-level collaboration.
2.  Use `{lintr}` to programmatically check and apply style guides to
    your project files.

## "Linting" your code

When you're working on your own and not needing to worry about anyone
seeing your code, there's a natural temptation to write your code like
you might write notes to yourself... scribbled and scrawled down
quickly. Even when you are working with others or try to write more
readable code, it still is very easy for "lint"[^linting-1] to
accumulate.

[^linting-1]: You know, the fluff you find in your pockets ðŸ˜

That's when "linters" (a type of "static code analysis" tool) become
very useful. A linter will scan your code for common mistakes or syntax
problems and list them out for you to fix. They don't directly fix them
for you, though we will show how to do that later in this session.
Linters are great when you are collaborating on a project with
collaborators who are not as experienced in writing code or who only
occasionally contribute so don't know the workflow culture of your
project. In this way, you might want to have automatic linting checks
(or styling) that are independent of you having to run them yourself.

The main package for doing that is the `{lintr}` package. We'll get a
chance to try out using `{renv}` since we need to install it to use it
in our project.

::: {.callout-note collapse="true"}
## Instructor note

Before continuing, ask the learners which type of dependency they think
`{lintr}` is or should be.
:::

```{r purl-only-renv-install-lintr}
#| eval: false
#| echo: false
#| purl: true
renv::install("lintr")
```

```{r add-lintr-as-dep}
#| eval: false
#| purl: true
use_package("lintr", "suggests")
```

```{r purl-only-update-desc-with-lintr}
#| eval: false
#| echo: false
#| purl: true
git_ci("DESCRIPTION", "Add lintr as a suggests dependency.")
```

Right now we don't have much code to "lint", but as we work through the
course, we will make use of `{lintr}` before committing to Git. So, how
do we use `{lintr}`? It's actually pretty easy, since there are only a
few functions that do the linting, even though there are dozens of
customizations you can make to `{lintr}`. The basic is to lint a single
file, don't run this though:

```{r lint-one-file}
#| eval: false
lintr::lint("R/functions.R")
```

A more encompassing function checks all files in the project (also,
don't run it yet, we'll do that when we have more code in the project):

```{r lint-project-dir}
#| eval: false
lintr::lint_dir()
```

This might be useful if you want to go over everything to check things.
But, if you are working on a single file and want to do some linting of
that, the better solution is to use the Command Palette (`Ctrl-Shift-P`)
and type "lint". The first option is to lint the file you are working
on. Let's open up the `data-raw/nmr-omics.R` script and try it out.
There *should* be at least one note that says:

``` markdown
Lines should not be more than 80 characters.
```

::: {.callout-note collapse="true"}
## Instructor note

If there is no note, check if the participants have the `# nolint` at
the end of the line with the second `download.file()`. To show what
happens, get them to remove it, run `lintr::lint_dir()` again, and then
bring it back.
:::

Keep in mind, that not all notes need to be changed. A linter is meant
to be used as a guideline, not as a rule. It helps you to follow best
practices and have some consistency between collaborators, but not to
stress about making sure all notes are addressed. We can suppress lint
warnings by adding a `# nolint` comment.

Since we will eventually connect our project Git repository to GitHub to
display a website, we can actually make use of a really powerful service
called [GitHub
Actions](https://docs.github.com/en/actions/learn-github-actions/understanding-github-actions)
to also run `{lintr}` there for us. But before we can do that, we need
to connect to GitHub. So before doing the next exercise, let's all
**commit the changes to the Git history**.

## Exercise: Connect your project to GitHub

> Time: \~25 minutes.

Let's complete these tasks to connect to GitHub.

1.  If you haven't yet, please create a
    [GitHub](https://github.com/join) account.

2.  Add the `{gitcreds}` package as a workflow dependency. You'll need
    it for the next item

    ``` r
    use_package("gitcreds", "suggests")
    ```

3.  Read through and complete the tasks in the [Connect to GitHub
    Guide]().

4.  Check your GitHub to make sure the project repository has been
    uploaded to it.

## Adding lintr to your GitHub

Now that your `AdvancedR3` project is connected and up on GitHub, we can
make use of the GitHub Actions. The RStudio team has a
[repository](https://github.com/r-lib/actions/tree/v2/examples#example-workflows)
that lists all R "Actions" that you can use to help streamline checks on
your project. Specifically, we want to add the one for `{lintr}`, which
we can add with:

```{r lintr-action}
#| eval: false
#| purl: true
use_github_action("lint-project")
```

```{r purl-only-add-lint-action}
#| eval: false
#| echo: false
#| purl: true
git_ci(".github", "Add lintr GitHub Action.")
```

This creates a file in `.github/workflows/lint-project.yaml`. In order
for GitHub to start using it, we need to **commit it to the Git history
and than push it to GitHub**.

Once you've done that, switch to your project on GitHub and click the
"Actions" tab at the top. Then click the "lint-project" link on the
sidebar. This lists the checks that `{lintr}` has done on your project
in an independent environment.

## Summary

-   Follow a style guide by using `{lintr}`. Combine with the Command
    Palette (`Ctrl-Shift-P`) to quickly run their functions on code you
    are actively working on.
